---
title: "neuroim2"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[![Build Status](https://img.shields.io/travis/bbuchsbaum/neuroim2/master.svg)](https://travis-ci.org/bbuchsbaum/neuroim2) [![codecov](https://codecov.io/github/bbuchsbaum/neuroim2/branch/master/graphs/badge.svg)](https://codecov.io/github/bbuchsbaum/neuroim2) 



The `neuroim2` package is currently in development. The goal of the project is to to provide basic functionality for working with neuroimaging data in R. `neuroim2` It is an effort to modify and upgrade the `neuroim` package (https://github.com/bbuchsbaum/neuroim). 


## Installation

You can install the development version of `neuroim2` (v. `0.1.0`) from Github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("bbuchsbaum/neuroim2")
```

## Usage

To read in a volumetric NIFTI formatted image, use the `read_vol` function:

```{r}
library(neuroim2)
fname <- system.file("extdata", "global_mask.nii", package="neuroim2")
vol <- read_vol(fname)
```

Now, `vol` can be treated as a three-dimensional array:

```{r}

v1 <- vol[1,1,1]
vol2 <- vol + vol
vol3 <- vol2 - vol
all(vol == vol3)
```

We can create a 4D image, by concatenating several 3D volumes:

```{r}

vec <- concat(vol, vol, vol2)
series1 <- vec[1,1,1,]
length(series1)
```

## Vignettes

See examples of use of `neuroim2` in the [vignettes](https://bbuchsbaum.github.io/neuroim2/articles/index.html).


