---
title: Choosing a Backend
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Choosing a Backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>

---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
suppressPackageStartupMessages(library(neuroim2))
```

# Why Backends Matter

`neuroim2` supports multiple representations for the same conceptual data:
3D volumes, 4D time-series, sparse masked data, memory-mapped arrays, and
file-backed access. Choosing the right representation is usually the biggest
practical decision you make up front.

This guide is the shortest path to that choice.

# Quick Rules

- Use `DenseNeuroVol` or `DenseNeuroVec` when the data fits comfortably in memory and you want the simplest behavior.
- Use `SparseNeuroVol` or `SparseNeuroVec` when a mask defines the stored voxel set, whether out-of-mask locations are outside the analysis support or simply implicit background zeros for your workflow.
- Use `MappedNeuroVec` when you want fast random access to an uncompressed NIfTI on disk without loading it all into memory.
- Use `FileBackedNeuroVec` when you want on-demand volume access from a file and can tolerate more I/O per access.
- Use `NeuroHyperVec` when the input is genuinely 5D.

# Decision Table

| Situation | Recommended backend | Why |
|:--|:--|:--|
| Small-to-moderate 3D image | `DenseNeuroVol` | simplest, lowest overhead |
| Small-to-moderate 4D fMRI series | `DenseNeuroVec` | easiest indexing and arithmetic |
| Masked analysis over a subset of voxels | `SparseNeuroVec` | stores only supported voxels |
| Large uncompressed NIfTI with random-access needs | `MappedNeuroVec` | memory mapping is efficient |
| Large file, on-demand volume access | `FileBackedNeuroVec` | no full materialization |
| 5D image | `NeuroHyperVec` | intended representation |

# Dense In-Memory Data

If the dataset fits in RAM, start here.

```{r}
file_name <- system.file("extdata", "global_mask_v4.nii", package = "neuroim2")
vec <- read_vec(file_name)

class(vec)
dim(vec)
```

This gives you the most predictable behavior for:

- arithmetic
- indexing
- resampling
- plotting
- converting to matrices or arrays

# Sparse Masked Data

If you only care about a subset of voxels, read the data through a mask and
keep the representation sparse.

```{r}
file_name <- system.file("extdata", "global_mask_v4.nii", package = "neuroim2")
mask_vol <- read_vol(file_name) > 0
svec <- read_vec(file_name, mask = mask_vol)

class(svec)
dim(svec)
sum(mask(svec))
```

Use this when:

- your workflow is defined on a brain mask or ROI support
- you want to avoid materializing out-of-mask voxels
- your downstream operations have clear semantics for absent voxels, such as outside-support values or implicit background zeros after densification

# Memory-Mapped Access

For large uncompressed NIfTI files, `mmap` mode gives efficient on-disk access.

```{r, eval = FALSE}
file_name <- system.file("extdata", "global_mask_v4.nii", package = "neuroim2")
mvec <- read_vec(file_name, mode = "mmap")

series(mvec, 1, 1, 1)
sub_vector(mvec, 1:5)
```

Use this when:

- the file is uncompressed
- you need repeated random access
- you want to avoid holding the full array in memory

# File-Backed Access

`filebacked` mode is the safest on-disk choice when you want deferred access
without requiring the full dense object in memory.

```{r, eval = FALSE}
file_name <- system.file("extdata", "global_mask_v4.nii", package = "neuroim2")
fbvec <- read_vec(file_name, mode = "filebacked")

fbvec[[1]]
sub_vector(fbvec, 1:3)
```

Use this when:

- you mostly work volume-by-volume
- you want an explicit on-disk representation
- you do not want to materialize the entire dataset

# 5D Data

When the input is 5D, use `read_image()` or `read_hyper_vec()`.

```{r, eval = FALSE}
img5d <- read_image("some_5d_image.nii.gz")
class(img5d)
```

# Recommended Defaults

If you are unsure:

1. Start with `read_vec()` or `read_vol()`.
2. Move to `SparseNeuroVec` only when a mask meaningfully reduces the support.
3. Move to `mmap` or `filebacked` when memory or I/O costs force the issue.

# Related Articles

- `vignette("Overview")` for general orientation
- `vignette("NeuroVector")` for 4D workflows
- `vignette("Resampling")` for spatial transforms
- `vignette("regionOfInterest")` for ROI and searchlight workflows
