---
title: "Working with Image Volumes"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Image Volumes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
```

## Reading a NIFTI formatted image volume

The way to read an volumetric image file is to use `read_vol`:


```{r}
    library(neuroim2)
    file_name <- system.file("extdata", "global_mask2.nii.gz", package="neuroim2")
    vol <- read_vol(file_name)
```

## Working with image volumes

Information about the geometry of the image volume is shown here:


```{r}
    print(vol)
```



`read_vol` returns an object of class `NeuroVol` object which extends an R `array` and has 3 dimensions (x,y,z).

```{r}
    class(vol)
    
    is.array(vol)
    
    dim(vol)
    
    vol[1,1,1]
    
    vol[64,64,24]
    
```
    
Arithmetic can be performed on images as if they were ordinary `array`s:


```{r}
    
    vol2 <- vol + vol
    sum(vol2) == 2 * sum(vol)
    
    vol3 <- vol2 - 2*vol
    all(vol3 == 0)
```

A numeric image volume can be converted to a binary image as follows:

```{r}
    
    vol2 <- as.logical(vol)
    class(vol2)
    print(vol2[1,1,1])
```

We can also create a `NeuroVol` instance from an `array` or `numeric` vector. First we consruct a standard R `array`:


```{r}
    x <- array(0, c(64,64,64))
```

Now we reate a `NeuroSpace` instance that describes the geometry of the image including, at minimum, its dimensions and voxel spacing.

```{r}
    bspace <- NeuroSpace(dim=c(64,64,64), spacing=c(1,1,1))
    vol <- NeuroVol(x, bspace)
    vol
```

We do not usually have to create `NeuroSpace` objects, because geometric information about an image is automatically determined from information stored in the image file header. Thus, `NeuroSpace` objects are usually copied from existing images using the `space` extractor function when needed:


```{r}
    vol2 <- NeuroVol((vol+1)*25, space(vol))
    max(vol2)
    space(vol2)
  
```

## Writing a NIFTI formatted image volume

When we're ready to write an image volume to disk, we use `write_vol`

```{r eval=FALSE}
    write_vol(vol2, "output.nii")
    
    ## adding a '.gz' extension results ina gzipped file.
    write_vol(vol2, "output.nii.gz")
```



