---
title: Large data
output:
  rmarkdown::html_vignette:
    css: albers.css
    includes:
      in_header: albers-header.html
vignette: |
  %\VignetteIndexEntry{Large data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
resource_files:
- albers.css
- albers.js
- albers-header.html
---

```{r setup, include = FALSE}
source("_common.R")
```

```{r load}
library(neuroim2)
```

Start dense. Move only when memory or I/O forces you to, and know what each move
costs — the backends differ enormously in *which* access pattern they make
cheap, and picking the wrong one is far worse than staying dense.

Everything below measures the same dataset, written to an uncompressed NIfTI so
that every backend can read it:

```{r data}
mask <- demo_mask()
brain <- as.mask(mask)
bold <- demo_bold(n_time = 60)

path <- tempfile(fileext = ".nii")
write_vec(bold, path)

dim(bold)
round(file.size(path) / 1e6, 1)
```

The file is uncompressed. Every backend can read `.nii.gz` too, but two of them
pay for it: memory mapping silently decompresses to a temporary uncompressed
copy, and file-backed access constructs happily and then fails at the first
read. Write uncompressed if you intend to map.

## What each backend costs

```{r backends}
dense <- read_vec(path)
sparse <- read_vec(path, mask = brain)
mapped <- read_vec(path, mode = "mmap")
filebacked <- read_vec(path, mode = "filebacked")

mb <- function(x) round(as.numeric(object.size(x)) / 1e6, 1)

data.frame(
  backend = c("DenseNeuroVec", "SparseNeuroVec", "MappedNeuroVec", "FileBackedNeuroVec"),
  memory_MB = c(mb(dense), mb(sparse), mb(mapped), mb(filebacked))
)
```

They all hold the same data:

```{r agreement}
ref <- as.numeric(series(dense, 32, 32, 12))

vapply(
  list(sparse = sparse, mapped = mapped, filebacked = filebacked),
  function(x) isTRUE(all.equal(as.numeric(series(x, 32, 32, 12)), ref)),
  logical(1)
)
```

## The number that actually decides it

Memory is the easy axis. The one that catches people is access *pattern*. Here
is the same work — twenty whole volumes, then twenty voxel time courses — on
each backend:

```{r timing}
per_call <- function(f, n) unname(system.time(for (i in seq_len(n)) f(i))["elapsed"] / n)

bench <- function(x, n = 20) {
  c(
    volumes = per_call(function(i) invisible(x[[i]]), n),
    series = per_call(function(i) series(x, 32, 32, 12), n)
  )
}

timings <- rbind(
  dense = bench(dense),
  sparse = bench(sparse),
  mapped = bench(mapped),
  filebacked = bench(filebacked, n = 5)
)

signif(timings, 2)
```

```{r timing-ratio}
round(timings["filebacked", "series"] / timings[c("dense", "sparse", "mapped"), "series"])
```

Seconds per call, and the ratio is computed rather than quoted, so it stays
honest on your machine. Fetching one voxel's time course from a
`FileBackedNeuroVec` costs three to four orders of magnitude more than it costs
anywhere else, because each query materialises the whole 4D
array as doubles before discarding all but one row — the disk read is a small
part of it. **Use it only when you genuinely work volume at a time.**
`MappedNeuroVec` is cheap on both patterns and is the better default.

## The ladder

**Dense** is the default. Predictable arithmetic, indexing and plotting, and
nothing to think about.

**Sparse** stores only the voxels a mask admits. Here the brain is 29% of the
grid, and the object is about a third the size:

```{r sparse}
c(in_mask = sum(brain), total = prod(dim(brain)))
c(dense_MB = mb(dense), sparse_MB = mb(sparse))
```

Dimensions are unchanged — sparsity is storage, not shape. Reach for it when a
mask defines your analysis anyway.

**Memory-mapped** leaves the data on disk and lets the operating system page it
in. Pointed at a `.nii.gz` it decompresses to a temporary uncompressed copy
first, so you pay the disk either way — better to store uncompressed
deliberately than to have a full copy appear in `tempdir()`.

```{r mmap}
class(mapped)[1]
dim(mapped)
```

`as_mmap()` converts an in-memory object, writing an uncompressed file if it has
to:

```{r as-mmap}
mmap_path <- tempfile(fileext = ".nii")
converted <- as_mmap(dense, file = mmap_path, overwrite = TRUE)

class(converted)[1]
isTRUE(all.equal(as.numeric(series(converted, 32, 32, 12)), ref))
```

Give it an explicit `file` when you can. Left to itself it writes into
`tempdir()` and you have a second full copy of the data with no obvious owner.

**BigNeuroVec** backs a masked series with an on-disk `bigstatsr` matrix, which
is the option for data larger than RAM that still needs voxelwise access:

```{r bigvec}
big <- read_vec(path, mode = "bigvec", mask = brain)

class(big)[1]
round(mb(big), 2)
```

### What object.size() does not see

The two on-disk backends report near-zero above, and that is `object.size()`
telling you the *R object* is a handle. It is not telling you the data is free:
mapped pages are charged to the process as you touch them, and `BigNeuroVec`
keeps a backing file — about 14 MB here — beside the original. Measure resident
memory and disk when the difference matters.

## Dropping to parcels

Every backend so far keeps one time course per voxel. The largest reduction
available comes from deciding you do not need them: a `ClusteredNeuroVec` stores
one series per *parcel*.

Parcels have to be spatially contiguous for any of this to mean anything.
Assigning labels at random would put all 100 parcels everywhere at once and
collapse their centroids onto the same point, so cluster them by position:

```{r clustered}
set.seed(1)
coords_in <- index_to_grid(mask, which(as.vector(brain)))
km <- kmeans(coords_in, centers = 100, iter.max = 30)

parcels <- ClusteredNeuroVol(brain, km$cluster)
cvec <- ClusteredNeuroVec(bold, parcels)

num_clusters(cvec)
dim(cvec)
round(apply(centroids(cvec), 2, sd), 1)
```

Those centroid standard deviations are voxels: the parcels are spread across the
brain, which is what makes centroid distances meaningful below.

It still behaves like a 4D image — `dim()` is unchanged and indexing works — but
only 100 time courses exist behind it:

```{r clustered-size}
dim(as.matrix(cvec, by = "cluster"))

c(dense_MB = mb(dense), clustered_MB = mb(cvec))
```

```{r clustered-access}
length(series(cvec, 32, 32, 12))
dim(cvec[, , , 1])

c(voxel_series = sum(brain), stored_series = num_clusters(cvec),
  fold = round(sum(brain) / num_clusters(cvec)))
```

Every voxel in a parcel returns that parcel's series: roughly a 300-fold
reduction in stored series here, and no within-parcel detail at all. For
atlas-based connectivity that is exactly right; for anything needing voxelwise
variation it is not.

One sharp edge: voxels outside the mask have no parcel, and `series()` returns
`NA` for them rather than zero or an error. Restrict to the mask before doing
arithmetic over the whole grid.

Cluster-level searchlights come with it, using centroid distances rather than
voxel neighbourhoods:

```{r cluster-searchlight}
windows <- cluster_searchlight_series(cvec, k = 3)

length(windows)
dim(values(windows[[1]]))
```

```{r cleanup, include = FALSE}
unlink(c(path, mmap_path, big@data$backingfile))
```

## Choosing

| Situation | Backend | Why |
|:--|:--|:--|
| It fits in memory | `DenseNeuroVec` | simplest, fastest, no caveats |
| A mask defines the analysis | `SparseNeuroVec` | stores only admitted voxels |
| Large uncompressed file, mixed access | `MappedNeuroVec` | cheap on both access patterns |
| Large file, strictly volume at a time | `FileBackedNeuroVec` | never for voxel series |
| Larger than RAM, needs voxel access | `BigNeuroVec` | on-disk matrix |
| Parcel-level analysis is enough | `ClusteredNeuroVec` | orders of magnitude fewer series |

If you are unsure: read with `read_vec()`, and change only when a measurement
tells you to. `object.size()` and `system.time()` will get you most of the way,
as long as you remember that the first does not see mapped pages or backing
files.

## Where to go next

- `vignette("volumes-and-vectors")` — the containers these all implement
- `vignette("regions-and-searchlights")` — parcels and searchlights in depth
- `?read_vec`, `?as_mmap`, `?ClusteredNeuroVec`, `?BigNeuroVec`
