---
title: Large data
output:
  rmarkdown::html_vignette:
    css: albers.css
    includes:
      in_header: albers-header.html
vignette: |
  %\VignetteIndexEntry{Large data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
resource_files:
- albers.css
- albers.js
- albers-header.html
---

```{r setup, include = FALSE}
source("_common.R")
```

```{r load}
library(neuroim2)
```

Start dense. Move only when memory or I/O forces you to, and know what each move
costs — the backends differ enormously in *which* access pattern they make
cheap, and picking the wrong one is far worse than staying dense.

Everything below measures the same dataset, written to an uncompressed NIfTI so
that every backend can read it:

```{r data}
bold <- demo_bold(n_time = 60)
brain <- as.mask(demo_mask() > 0)

path <- tempfile(fileext = ".nii")
write_vec(bold, path)

c(dim = dim(bold), file_MB = round(file.size(path) / 1e6, 1))
```

## What each backend costs

```{r backends}
dense <- read_vec(path)
sparse <- read_vec(path, mask = brain)
mapped <- read_vec(path, mode = "mmap")
filebacked <- read_vec(path, mode = "filebacked")

mb <- function(x) round(as.numeric(object.size(x)) / 1e6, 1)

data.frame(
  backend = c("DenseNeuroVec", "SparseNeuroVec", "MappedNeuroVec", "FileBackedNeuroVec"),
  memory_MB = c(mb(dense), mb(sparse), mb(mapped), mb(filebacked))
)
```

They all hold the same data:

```{r agreement}
ref <- as.numeric(series(dense, 32, 32, 12))

vapply(
  list(sparse = sparse, mapped = mapped, filebacked = filebacked),
  function(x) isTRUE(all.equal(as.numeric(series(x, 32, 32, 12)), ref)),
  logical(1)
)
```

## The number that actually decides it

Memory is the easy axis. The one that catches people is access *pattern*. Here
is the same work — twenty whole volumes, then twenty voxel time courses — on
each backend:

```{r timing}
time_it <- function(expr) round(system.time(expr)["elapsed"], 2)

set.seed(1)
idx <- sample(which(as.vector(brain)), 20)
g <- index_to_grid(demo_mask(), idx)

bench <- function(x) {
  c(
    volumes = time_it(for (i in 1:20) invisible(x[[i]])),
    series = time_it(for (i in 1:20) series(x, g[i, 1], g[i, 2], g[i, 3]))
  )
}

round(rbind(
  dense = bench(dense),
  sparse = bench(sparse),
  mapped = bench(mapped),
  filebacked = bench(filebacked)
), 2)
```

`FileBackedNeuroVec` reads a whole volume to answer any question. Asking it for
one voxel's time course means reading the entire file once per voxel, which is
why that cell is thousands of times slower than the others rather than a little
slower. **Use it only when you genuinely work volume at a time.**
`MappedNeuroVec` is cheap on both patterns and is the better default whenever the
file is uncompressed.

## The ladder

**Dense** is the default. Predictable arithmetic, indexing and plotting, and
nothing to think about.

**Sparse** stores only the voxels a mask admits. Here the brain is 29% of the
grid, and the object is about a third the size:

```{r sparse}
c(in_mask = sum(brain), total = prod(dim(brain)))
c(dense_MB = mb(dense), sparse_MB = mb(sparse))
```

Dimensions are unchanged — sparsity is storage, not shape. Reach for it when a
mask defines your analysis anyway.

**Memory-mapped** leaves the data on disk and lets the operating system page it
in. It needs an *uncompressed* file, which is the one real constraint: a
`.nii.gz` cannot be mapped.

```{r mmap}
class(mapped)[1]
dim(mapped)
```

`as_mmap()` converts an in-memory object, writing an uncompressed file if it has
to:

```{r as-mmap}
converted <- as_mmap(dense)

class(converted)[1]
isTRUE(all.equal(as.numeric(series(converted, 32, 32, 12)), ref))
```

**BigNeuroVec** backs a masked series with an on-disk `bigstatsr` matrix, which
is the option for data larger than RAM that still needs voxelwise access:

```{r bigvec}
big <- read_vec(path, mode = "bigvec", mask = brain)

class(big)[1]
mb(big)
```

## Dropping to parcels

Every backend so far keeps one time course per voxel. The largest reduction
available comes from deciding you do not need them: a `ClusteredNeuroVec` stores
one series per *parcel*.

```{r clustered}
set.seed(1)
parcels <- ClusteredNeuroVol(brain, sample(1:100, sum(brain), replace = TRUE))
cvec <- ClusteredNeuroVec(bold, parcels)

num_clusters(cvec)
dim(cvec)
```

It still behaves like a 4D image — `dim()` is unchanged and indexing works — but
only 100 time courses exist behind it:

```{r clustered-size}
dim(as.matrix(cvec, by = "cluster"))

c(dense_MB = mb(dense), clustered_MB = mb(cvec))
```

```{r clustered-access}
length(series(cvec, 32, 32, 12))
dim(cvec[, , , 1])
```

Every voxel in a parcel returns that parcel's series, which is the trade: a
100-fold reduction in stored series, and no within-parcel detail at all. For
atlas-based connectivity this is exactly right; for anything that needs
voxelwise variation it is not.

Cluster-level searchlights come with it, using centroid distances rather than
voxel neighbourhoods:

```{r cluster-searchlight}
windows <- cluster_searchlight_series(cvec, k = 3)

length(windows)
dim(values(windows[[1]]))
```

```{r cleanup, include = FALSE}
unlink(path)
```

## Choosing

| Situation | Backend | Why |
|:--|:--|:--|
| It fits in memory | `DenseNeuroVec` | simplest, fastest, no caveats |
| A mask defines the analysis | `SparseNeuroVec` | stores only admitted voxels |
| Large uncompressed file, mixed access | `MappedNeuroVec` | cheap on both access patterns |
| Large file, strictly volume at a time | `FileBackedNeuroVec` | never for voxel series |
| Larger than RAM, needs voxel access | `BigNeuroVec` | on-disk matrix |
| Parcel-level analysis is enough | `ClusteredNeuroVec` | orders of magnitude fewer series |
| Genuinely 5D input | `NeuroHyperVec` | the intended representation |

If you are unsure: read with `read_vec()`, and change only when a measurement
tells you to. `object.size()` and `system.time()`, as used above, are the whole
toolkit.

## Where to go next

- `vignette("volumes-and-vectors")` — the containers these all implement
- `vignette("regions-and-searchlights")` — parcels and searchlights in depth
- `?read_vec`, `?as_mmap`, `?ClusteredNeuroVec`, `?BigNeuroVec`
