---
title: 'neuroim2: a tour'
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{neuroim2: a tour}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>
---

```{r setup, include = FALSE}
source("_common.R")
```

```{r load}
library(neuroim2)
```

## Look at an image

```{r anatomy, fig.cap = "plot() on a NeuroVol: a 3 x 3 montage of evenly spaced axial slices.", fig.alt = "Nine axial slices of an anatomical brain image.", fig.height = 6.5}
anat <- read_vol(system.file("extdata", "mni_downsampled.nii.gz", package = "neuroim2"))
plot(anat)
```

That is a brain, read from a NIfTI file in one line and drawn in another. It is
also an ordinary R array:

```{r array-like}
dim(anat)
anat[24, 28, 24]
mean(anat > 0)
```

Indexing, arithmetic and comparison all work the way you expect, and the result
of `anat > 0` is still an image rather than a bare logical array. Everything in
this package is built on that idea: objects that behave like arrays but never
forget their geometry.

## Geometry is the part that matters

An image is not just numbers, it is numbers *somewhere*. That somewhere is a
`NeuroSpace`, attached to every object, and it is what separates neuroimaging
from array programming.

Here is the functional dataset this tour analyses — a brain mask on the grid an
EPI run was acquired on. Anatomical and functional images are separate
acquisitions with separate grids, which is why we now switch images.

```{r mask}
mask <- read_vol(system.file("extdata", "global_mask2.nii.gz", package = "neuroim2"))

space(mask)
```

The space records voxel size in millimetres, where the grid sits in the scanner,
and which anatomical direction each axis runs:

```{r geometry}
spacing(mask)
affine_to_axcodes(trans(mask))
```

So `spacing()` tells you these voxels are 3.5 x 3.5 x 3.7 mm, and the axis codes
say the first axis increases towards the **L**eft, the second towards the
**A**nterior, the third towards the **S**uperior. With that, the package can
answer the question you actually care about: *which voxel sits at this
anatomical location?*

```{r coord}
coord_to_grid(mask, c(-34, -28, 10))
```

Left auditory cortex, in millimetres, becomes a position on this grid. The
answer is fractional because a millimetre coordinate rarely lands on a voxel
centre. `grid_to_coord()` goes back the other way:

```{r coord-back}
grid_to_coord(mask, matrix(c(42.714, 24, 16.203), nrow = 1))
```

Keep the fractional position if you are going to interpolate, and round it only
when you need an integer index — rounding moves you up to half a voxel, which on
this grid is nearly 2 mm:

```{r coord-round}
grid_to_coord(mask, matrix(c(43, 24, 16), nrow = 1))
```

## Add a time axis

Functional data has a fourth dimension. `read_vec()` reads 4D files; here we
simulate a run instead, so that the time series in this article have the
temporal structure real BOLD data has.

```{r bold}
bold <- simulate_fmri(mask, n_time = 60, seed = 1)

dim(bold)
```

A `NeuroVec` is 60 volumes sharing one spatial frame. Pull the time course of a
single voxel with `series()`:

```{r series, fig.cap = "One voxel's simulated BOLD time course.", fig.alt = "Line plot of a single voxel time series over 60 scans."}
ts <- series(bold, 43, 24, 16)

plot(ts,
  type = "l", xlab = "scan", ylab = "signal",
  main = "Voxel (43, 24, 16)"
)
```

## Summarise a region

Single voxels are noisy, so analyses usually work over a region. Put a 10 mm
sphere at the coordinate from earlier and extract every time course inside it.
`nonzero = TRUE` keeps only voxels that are actually in the mask.

```{r roi}
roi <- spherical_roi(mask, c(43, 24, 16), radius = 10, nonzero = TRUE)
length(roi)

roi_ts <- series_roi(bold, roi)
dim(values(roi_ts))
```

`values()` gives scans by voxels, so averaging across columns gives the region's
mean time course:

```{r roi-mean, fig.cap = "Averaging over 85 voxels suppresses noise the single voxel had.", fig.alt = "Line plot of the ROI mean time series, visibly smoother than the single voxel."}
roi_mean <- rowMeans(values(roi_ts))

plot(roi_mean,
  type = "l", xlab = "scan", ylab = "signal",
  main = "ROI mean, 85 voxels"
)
```

That is the shape of most work in this package: **define a spatial support,
extract values from it, reduce them to something smaller.**

## Make a result and write it out

Reductions run the other way too — from a 4D series down to one volume. Here is
each voxel's temporal standard deviation, a standard quality-control map:

```{r sdmap, fig.cap = "Temporal standard deviation, computed per voxel and returned as a NeuroVol.", fig.alt = "Nine axial slices of a temporal standard deviation map.", fig.height = 6.5}
sd_map <- NeuroVol(
  apply(as.matrix(bold), 1, sd),
  drop_dim(space(bold))
)

plot(sd_map)
```

Two details carry the geometry through: `as.matrix()` flattens a `NeuroVec` to
voxels-by-time, and `drop_dim()` takes the 4D space down to the matching 3D one.
The result is a proper `NeuroVol`, so writing it produces a NIfTI any other tool
can read.

```{r write}
out <- tempfile(fileext = ".nii.gz")
write_vol(sd_map, out)

file.exists(out)
```

```{r cleanup, include = FALSE}
unlink(out)
```

## Where to go next

You have now read images, inspected their geometry, moved between millimetres
and voxels, extracted a region's time course, reduced a series to a map, and
written it back to disk. Three articles take that further:

- `vignette("spaces-and-coordinates")` — the affine, orientation codes and
  conversions in full. Read this one next; everything else assumes it.
- `vignette("volumes-and-vectors")` — the container types and how to build,
  slice, combine and convert them.
- `vignette("regions-and-searchlights")` — ROIs, parcels and searchlights, and
  the split–map–reduce patterns built on them.

Reference pages worth knowing early: `?read_vol`, `?NeuroSpace`,
`?spherical_roi`, `?series`.
