---
title: "Working with 4D NeuroVectors (NeuroVec)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with 4D NeuroVectors (NeuroVec)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
library(purrr)
library(assertthat)
library(neuroim2)
options(mc.cores=1)
```

## Working with neuroimaging time-series data

The `neuroim2` package contains data structures and functions for reading, accessing, and processing 4-dimensional neuroimaging data. 

In this vignette, we’ll introduce `NeuroVec` (4D images) and related helpers you’ll use most often:

- Read/write on-disk images (`read_vec`, `write_vec`)
- Spatial metadata via `NeuroSpace` (dimensions, spacing, origin)
- Voxel- and ROI-based access (`series`, `series_roi`, `coords`)
- Common transforms (z-scoring with `scale_series`, concatenation with `concat`)
- Dense vs. sparse representations (`DenseNeuroVec`, `SparseNeuroVec`, `as.sparse`)

### Reading a four-dimensional NifTI image with read_vec

Here we read in an example image. This file is 3D (single timepoint) in the
package data; the interface is the same for 4D images.

```{r}
      library(purrr)
      library(ggplot2)
      file_name <- system.file("extdata", "global_mask_v4.nii", package="neuroim2")
      vec <- read_vec(file_name)
      dim(vec)
      vec
```

Now imagine we have a set of images. We can read multiple files with `read_vec`.
Passing multiple paths returns a `NeuroVecSeq` (a sequence of vectors) rather than a
single concatenated 4D vector.


```{r}
    
      file_name <- system.file("extdata", "global_mask_v4.nii", package="neuroim2")
      vec_multi <- read_vec(c(file_name, file_name, file_name))
      dim(vec_multi)
      
      vec2 <- read_vec(rep(file_name, 10))
      vec2
```

To extract a subset of volumes from a 4D vector, we’ll first create a small
time series by concatenating a 3D volume:

```{r}
      
      # Build a 4D vector with 10 timepoints by concatenation
      vol <- read_vol(file_name)
      vec_t <- concat(vol, vol, vol, vol, vol, vol, vol, vol, vol, vol)
      
      # Extract a subset of volumes
      vec_1_6 <- sub_vector(vec_t, 1:6)
      dim(vec_1_6)
      vec_1_6
```

### Extracting time-series data using the `series` and `series_roi` functions

To get the time-series at voxel (1,1,1) we can use the `series` function:

```{r}
      
      series(vec_1_6, 1,1,1)
```

We can extract a 4d region of interest with the `series_roi` as follows:


```{r}
      file_name <- system.file("extdata", "global_mask_v4.nii", package="neuroim2")
      vol <- read_vol(file_name)
      roi <- spherical_roi(vol, c(12,12,12), radius=8)
      rvec1 <- series_roi(vec, roi)
      
      ## or alternatively as a pipeline
      rvec2 <- read_vol(file_name) %>% spherical_roi(c(12,12,12), radius=8) %>% series_roi(vec,.)
      rvec2
      
      ## we can extract the ROI values with the `values` method.
      assertthat::assert_that(all(values(rvec1) == values(rvec2)))
      assertthat::assert_that(all(coords(rvec1) == coords(rvec2)))
      
```

We can also extract an ROI using 1d indices:

```{r}

r1 <- series_roi(vec, 1:100)
r1
```

Or we can extract a plain matrix using the `series` function:

```{r}
r2 <- series(vec, 1:100)
dim(r2)
```

We can also use coordinate indexing using voxel coordinates. First we load a binary mask with the same spatial dimensions as our NeuroVec:

```{r}
mask <- read_vol(system.file("extdata", "global_mask_v4.nii", package="neuroim2"))
```

Now we convert indices to voxels and extract a matrix of values at the specified locations:

```{r}
vox <- index_to_grid(mask, 1:100)

r3 <- series(vec, vox)
dim(r3)
```

And the same using `series_roi`:

```{r}
r4 <- series_roi(vec,vox)
r4
```


### Inspecting spatial metadata with NeuroSpace

Every `NeuroVec` carries a `NeuroSpace` describing its geometry.

```{r}
sp <- space(vec_t)
sp                   # dimensions, spacing, origin, axes
dim(vec_t)           # 4D dims: X × Y × Z × T
spacing(vec_t)       # voxel spacing (mm)
origin(vec_t)        # image origin
ndim(vec_t)          # == 4 for time series
```

The default mask for dense vectors is “all voxels are valid”:

```{r}
m <- mask(vec_t)
m
```

### Creating a NeuroVec from an in-memory array

You can build a `NeuroVec` directly from arrays or matrices:

```{r}
set.seed(1)
dims <- c(24, 24, 24, 5)
arr  <- array(rnorm(prod(dims)), dims)
sp4  <- NeuroSpace(dims, spacing = c(2,2,2))
dvec <- NeuroVec(arr, sp4)
dim(dvec)
```

You can also start from a matrix (voxels × time or time × voxels) using `DenseNeuroVec`:

```{r}
mat  <- matrix(rnorm(prod(dims)), nrow = prod(dims[1:3]), ncol = dims[4])
dvec2 <- DenseNeuroVec(mat, sp4)
all.equal(dim(dvec), dim(dvec2))
```

### Time-series transforms: z-scoring and summary volumes

Z-score each voxel’s time-series (center and scale across time):

```{r}
vec_z <- scale_series(dvec, center = TRUE, scale = TRUE)
dim(vec_z)
```

Compute a mean volume across time and return a 3D `NeuroVol`:

```{r}
M      <- as.matrix(dvec)              # voxels × time
vmean  <- rowMeans(M)                  # per-voxel mean
mean3d <- NeuroVol(vmean, drop_dim(space(dvec)))
mean3d
```

### Concatenating along time

Append time points by concatenating vectors or volumes:

```{r}
dvec_more <- concat(dvec, dvec)        # doubles the time dimension
dim(dvec_more)
```

### Dense ↔ sparse workflows

Sparse representations store only voxels within a mask. This is handy for large ROIs or brain masks.

```{r}
# Build a random mask and convert a dense vec to sparse
mask_arr <- array(runif(prod(dims[1:3])) > 0.7, dims[1:3])
mask_vol <- LogicalNeuroVol(mask_arr, drop_dim(sp4))

svec <- as.sparse(dvec, mask_vol)     # SparseNeuroVec with explicit mask
svec                                 # note the stored mask and cardinality

# Convert back to dense if needed
dense_again <- as.dense(svec)
all.equal(dim(dense_again), dim(dvec))
```

Tip: For file-backed or memory-mapped vectors, convert to `DenseNeuroVec` via a matrix if you need dense-only operations:

```{r}
dv_dense <- DenseNeuroVec(as.matrix(vec), space(vec))
```

### Writing vectors to disk

You can write `NeuroVec` and `NeuroVol` objects as NIfTI files:

```{r}
tmp_vec <- tempfile(fileext = ".nii.gz")
write_vec(vec_1_6, tmp_vec)
file.exists(tmp_vec)
unlink(tmp_vec)
```

### Putting it together with an ROI

Combine ROI extraction with time-series transforms:

```{r}
vol3d <- read_vol(system.file("extdata", "global_mask_v4.nii", package="neuroim2"))
roi   <- spherical_roi(vol3d, c(12,12,12), radius = 6)
rts   <- series_roi(vec, roi)          # ROIVec (T × N with coords)

# z-score each column (voxel) across time, then average within ROI
mat_roi  <- values(rts)                # T × N
mat_z    <- base::scale(mat_roi, center=TRUE, scale=TRUE)
roi_mean <- rowMeans(mat_z)
length(roi_mean)                       # matches time dimension
```

That’s the core workflow for 4D data in `neuroim2`: load or create a `NeuroVec`, inspect its `NeuroSpace`, access time-series via voxels or ROIs, apply simple transforms, and optionally move between dense and sparse representations.

