---
title: "Pipelines"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pipelines}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
devtools::load_all()
library(assertthat)
```

Pipelining operations using a functional approach
===================

The `neuroim2` packages provides a set of functions that allows one to split image data in various ways to processing data split into parts. By breaking a dataset up into pieces, we can also more easily parallelize certain operations. 

## Splitting an image into connected components

First we load in an example volume, assign it random values, and find its connected components with a threshold of .9

```{r}
      library(purrr)
      library(ggplot2)
      file_name <- system.file("extdata", "global_mask.nii", package="neuroim2")
      vol <- read_vol(file_name)
      mask.idx <- which(vol>0)
      vol[mask.idx] <- runif(length(mask.idx))
      comp <- conn_comp(vol, threshold=.9)
      
      plot(comp$index, zlevels=seq(1,25,by=3), cmap=rainbow(255))
```

Now we want to find the average value in each of the connected components using the `split_clusters` function. Since `conn_comp` returns a `ClusteredNeuroVol` containing the cluster indices, we use that to split the original volume into a list of `ROIVol`s and compute the mean over each one.

```{r}

mvals <- vol %>% split_clusters(comp$index) %>% map_dbl( ~ mean(.))

```
    
Suppose we want to compute the local standard deviation within a 4mm radius of each voxel. We can use the `searchlight` function to construct a list of spherical ROIs centered on every voxel in the input set.

```{r}

sdvol <- vol %>% searchlight(radius=5, eager=TRUE) %>% map_dbl( ~ sd(.)) %>% NeuroVol(space=space(vol), indices=which(vol!=0))
plot(sdvol, cmap=rainbow(255))

```

Another thing we might to is compute the k nearest neighbors in each searchlight and replace the center voxel with the average intensity of its neighbors:

```{r}

k <- 12
knnfvol <- vol %>% searchlight(radius=12, eager=TRUE) %>% map_dbl(function(x) {
  ind <- order((x[x@center_index] - x)^2)[1:k]
  mean(x[ind])
}) %>% NeuroVol(space=space(vol), indices=which(vol!=0))

plot(knnfvol, cmap=rainbow(255), zlevels=seq(1,25,by=3))

```
If we only need access to the searchlight coordinates (in voxel space), we can use the `searchlight_coords` function. Here, we simply replace the center voxel with the average of its neighbors in searchlight space:

```{r}

avgvol <- vol %>% searchlight_coords(radius=12) %>% map_dbl(function(x) {
  vals <- vol[x]
  mean(vals[vals!=0])
}) %>% NeuroVol(space=space(vol), indices=which(vol!=0))

plot(avgvol, cmap=rainbow(255), zlevels=seq(1,25,by=3))

```
## Mapping a function over each volume of a `NeuroVec`

```{r}

vec <- concat(vol,vol,vol,vol,vol)
mean_vec <- vec %>% vols %>% map_dbl(~ mean(.))
sd_vec <- vec %>% vols %>% map_dbl(~ sd(.))
assert_that(length(mean_vec) == dim(vec)[4])
assert_that(length(sd_vec) == dim(vec)[4])

```

## Mapping a function over each vector of a NeuroVec

```{r}
vec <- concat(vol,vol,vol,vol,vol)
f <- function(x) { browser() }
mean_vol <- vec %>% vectors %>% map_dbl(~ mean(.)) %>% NeuroVec(., space=space(vol))
assert_that(all(dim(mean_vol) == dim(vol)))
```


