---
title: 'Pipelines: Split, Map, and Reduce'
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: '%\VignetteIndexEntry{Pipelines: Split, Map, and Reduce} %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}'
params:
  family: red
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',()=>document.body.classList.add('palette-red'));</script>

---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>", fig.alt = 'Plot output')
library(purrr)
library(assertthat)
library(neuroim2)
options(mc.cores=1)
```

Pipelining operations using a functional approach
===================

The `neuroim2` packages provides a set of functions that allows one to split image data in various ways to processing data split into parts. By breaking a dataset up into pieces, we can also more easily parallelize certain operations. 

## Splitting an image into connected components

First we load in an example volume, assign it random values, and find its connected components with a threshold of .9

```{r}
      library(purrr)
      library(ggplot2)
      file_name <- system.file("extdata", "global_mask_v4.nii", package="neuroim2")
      vol <- read_vol(file_name)
      mask.idx <- which(vol>0)
      
      vol2 <- vol
      vol2[mask.idx] <- runif(length(mask.idx))
      comp <- conn_comp(vol2, threshold=.8)
      
      plot(comp$index, zlevels=seq(1,25,by=3), cmap=rainbow(255))
```

Now we want to find the average value in each of the connected components using the `split_clusters` function. Since `conn_comp` returns a `ClusteredNeuroVol` containing the cluster indices, we use that to split the original volume into a list of `ROIVol`s and compute the mean over each one.

```{r}
mvals <- vol2 %>% split_clusters(comp$index) %>% map_dbl( ~ mean(.))
```
    
Suppose we want to compute the local standard deviation within a 4mm radius of each voxel. We can use the `searchlight` function to construct a list of spherical ROIs centered on every voxel in the input set.

```{r}
sdvol <- vol %>% searchlight(radius=5, eager=TRUE) %>% map_dbl( ~ sd(values(.))) 
sdvol <- NeuroVol(sdvol, space=space(vol), indices=which(vol!=0))
plot(sdvol, cmap=rainbow(255))
```

Another thing we might to is compute the k nearest neighbors in each searchlight and replace the center voxel with the average intensity of its neighbors:

```{r}
k <- 12
knnfvol <- vol2 %>% searchlight(radius=6, eager=TRUE) %>% map_dbl(function(x) {
  # Just compute mean of all values in the searchlight
  mean(values(x))
}) %>% NeuroVol(space=space(vol), indices=which(vol!=0))
plot(knnfvol, cmap=rainbow(255))
```

If we only need access to the searchlight coordinates (in voxel space), we can use the `searchlight_coords` function. Here, we simply replace the center voxel with the average of its neighbors in searchlight space:

```{r}
avgvol <- vol %>% searchlight_coords(radius=12, nonzero=TRUE) %>% map_dbl(function(x) {
  vals <- vol[x]
  mean(vals[vals!=0])
}) %>% NeuroVol(space=space(vol), indices=which(vol!=0))
plot(avgvol, cmap=rainbow(2), zlevels=seq(1,25,by=3))
```

## Mapping a function over every slice of a `NeuroVol`

Suppose we want to split up an image volume by slice and apply a function to each slice. We can use the `slices` function to achieve this as follows:

```{r}
slice_means <- vol %>% slices %>% map_dbl(~ mean(.))
plot(slice_means, type='l', ylab="mean intensity", xlab="slice number")
```

## Mapping a function over each volume of a `NeuroVec` object

```{r}
vec <- concat(vol,vol,vol,vol,vol)
vec
mean_vec <- vec %>% vols %>% map_dbl(~ mean(.))
sd_vec <- vec %>% vols %>% map_dbl(~ sd(.))
assert_that(length(mean_vec) == dim(vec)[4])
assert_that(length(sd_vec) == dim(vec)[4])
```

## Mapping a function over each vector of a `NeuroVec` object

```{r}
vec <- concat(vol,vol,vol,vol,vol)
vec
mean_vol <- vec %>% vectors() %>% map_dbl(~ mean(.)) %>% NeuroVol(., space=space(vol))
assert_that(all(dim(mean_vol) == dim(vol)))
```
