---
title: Reading and writing images
output:
  rmarkdown::html_vignette:
    css: albers.css
    includes:
      in_header:
      - albers-header.inc
      - albers-header.html
    toc: yes
    toc_depth: 2.0
resource_files:
- albers.css
- albers.js
- albers-header.inc
- albers-header.html
- fonts
params:
  family: red
  preset: interaction

vignette: |
  %\VignetteIndexEntry{Reading and writing images}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
source("_common.R")
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

```{r load}
library(neuroim2)
```

Every analysis starts and ends at a file. This article covers the readers and
writers, how to inspect a header before committing to loading anything, and the
one header problem that produces wrong results without producing an error.

```{r paths}
mask_file <- system.file("extdata", "global_mask2.nii.gz", package = "neuroim2")
series_file <- system.file("extdata", "global_mask_v4.nii", package = "neuroim2")
anat_file <- system.file("extdata", "mni_downsampled.nii.gz", package = "neuroim2")
```

## Choosing a reader

| Function | Returns | Use for |
|:--|:--|:--|
| `read_vol()` | `NeuroVol` | a 3D image, or one volume of a 4D one |
| `read_vec()` | `NeuroVec` | a 4D series |
| `read_image()` | whichever fits | unknown dimensionality |
| `read_vol_list()` | `NeuroVec` | many 3D files as one series |
| `read_header()` | `FileMetaInfo` | metadata without the data |

```{r readers}
vol <- read_vol(mask_file)
vec <- read_vec(series_file)

dim(vol)
dim(vec)
```

Compression is handled by extension: `.nii` and `.nii.gz` both work on read and
on write. `read_vol()` on a 4D file returns its first volume, and `index` picks
a different one:

```{r read-index}
dim(read_vol(series_file, index = 3))
```

Use `read_image()` when you do not know what a file holds:

```{r read-image}
class(read_image(series_file))[1]
```

Reading several files gives you a 4D object either way, but not the same one.
`read_vol_list()` builds a single contiguous series; `read_vec()` on a vector of
paths keeps the runs as separable segments:

```{r multi}
class(read_vol_list(c(mask_file, mask_file)))[1]
class(read_vec(c(series_file, series_file)))[1]

dim(read_vec(c(series_file, series_file)))
```

Neither returns a list, despite the name — `length()` on either counts volumes.

## Looking before you load

`read_header()` reads metadata only. On a large file this costs nothing and
tells you what you are about to commit to:

```{r header}
hdr <- read_header(mask_file)

dim(hdr)
hdr@data_type
hdr@spacing
```

`dim()` and `trans()` work on a header, which is the quickest way to see the
affine neuroim2 will actually use without loading a byte of data:

```{r header-trans}
trans(hdr)
```

`spacing()`, `origin()` and `space()` are not defined for headers — read the
slots (`hdr@spacing`, `hdr@origin`) instead. The full NIfTI header is a plain
list in the `header` slot, so anything the slots omit is still reachable:

```{r header-slot}
hdr@header$datatype
hdr@header$encoding
hdr@header$vox_offset
```

## The two affines

A NIfTI file can carry **two** spatial transforms: the `qform` (a rigid
transform stored as a quaternion) and the `sform` (a general affine). Each has a
code saying whether it is set. The standard says use the sform when
`sform_code > 0`, otherwise the qform.

On a well-formed file the two agree:

```{r affines-good}
c(qform_code = hdr@header$qform_code, sform_code = hdr@header$sform_code)

all.equal(hdr@header$qform, hdr@header$sform)
```

They do not always agree. The anatomical image shipped with this package is a
real example — its qform describes 4 mm voxels and its sform describes 1 mm
ones:

```{r affines-bad}
anat_hdr <- read_header(anat_file)

anat_hdr@header$pixdim[2:4]
anat_hdr@header$qform
anat_hdr@header$sform
```

Both codes are set, so the sform wins and the image reports 1 mm voxels even
though `pixdim` says otherwise:

```{r affines-consequence}
spacing(read_vol(anat_file))
```

Nothing errors. The image loads and plots correctly, and every **distance,
volume and ROI radius** derived from it is wrong by a factor of about four. The
coordinates themselves are not scaled by four — both affines share a translation,
so they agree exactly at voxel `(1, 1, 1)` and diverge by a growing offset that
reaches roughly 140 mm at the far corner:

```{r affines-divergence}
far <- matrix(dim(read_vol(anat_file)), nrow = 1)

sform_xyz <- grid_to_coord(space(read_vol(anat_file)), far)
qform_xyz <- (anat_hdr@header$qform %*% c(far - 1, 1))[1:3]

rbind(sform = as.vector(sform_xyz), qform = qform_xyz)
```

**This is the failure mode to know about**, because no amount of care downstream
will catch it. Check when a file first arrives:

```{r affine-check}
affines_agree <- function(file) {
  h <- read_header(file)
  qc <- h@header$qform_code
  sc <- h@header$sform_code
  if (is.null(qc) || is.null(sc) || qc <= 0 || sc <= 0) return(NA)
  isTRUE(all.equal(h@header$qform, h@header$sform, tolerance = 1e-4))
}

c(mask = affines_agree(mask_file), anat = affines_agree(anat_file))
```

The `is.null()` guard matters: not every format carries these fields, and
returning `NA` for "cannot tell" is better than erroring on an AFNI file.

If a file is affected and you trust its `pixdim`, rebuild the geometry:

```{r repair}
bad <- read_vol(anat_file)
fixed <- NeuroVol(
  as.array(bad),
  NeuroSpace(dim(bad),
    spacing = anat_hdr@header$pixdim[2:4],
    origin = anat_hdr@header$qform[1:3, 4],
    axes = space(bad)@axes
  )
)

spacing(fixed)
```

## Writing

`write_vol()` and `write_vec()` take an object and a path. The extension decides
the container and the compression; an unrecognised or absent extension gives
uncompressed NIfTI.

```{r write}
out_nii <- tempfile(fileext = ".nii")
out_gz <- tempfile(fileext = ".nii.gz")

write_vol(vol, out_nii)
write_vol(vol, out_gz)

c(plain = file.size(out_nii), gzipped = file.size(out_gz))
```

That plain file is 400 kB for a binary mask, because `write_vol()` promotes to
`FLOAT` by default. Pass `data_type` when you know better:

```{r datatype}
out_byte <- tempfile(fileext = ".nii")
write_vol(vol, out_byte, data_type = "UBYTE")

c(float = file.size(out_nii), ubyte = file.size(out_byte))
read_header(out_byte)@data_type
```

Valid types are `UNKNOWN`, `BINARY`, `UBYTE`, `SHORT`, `INT`, `FLOAT` and
`DOUBLE`. A round trip preserves both data and geometry either way:

```{r roundtrip}
back <- read_vol(out_gz)

all.equal(as.array(back), as.array(vol))
identical(trans(space(back)), trans(space(vol)))
```

4D objects work the same way:

```{r write-vec}
out_vec <- tempfile(fileext = ".nii.gz")
write_vec(sub_vector(vec, 1:2), out_vec)

dim(read_vec(out_vec))
```

### A precision surprise that is not about files

Affines round to seven significant figures, and spacing and origin to six — but
this happens in `NeuroSpace()`, when the object is built, not when it is written:

```{r precision}
sp <- NeuroSpace(c(4L, 4L, 4L), spacing = c(1 / 3, 0.123456789012, pi))

spacing(sp)
```

No file was involved. Because both sides of a round trip get the same rounding,
`identical()` on a re-read affine does succeed, as above. What you cannot rely on
is an affine you constructed matching the full-precision numbers you passed in.

```{r cleanup, include = FALSE}
unlink(c(out_nii, out_gz, out_byte, out_vec))
```

## AFNI files

AFNI `HEAD`/`BRIK` pairs are read by the same functions — pass the `.HEAD` file
and the matching `.BRIK` is found beside it. `read_header()` returns an
`AFNIMetaInfo` whose `header` slot holds the AFNI attribute list:

```{r afni, eval = FALSE}
vol <- read_vol("subject_anat+orig.HEAD")
names(read_header("subject_anat+orig.HEAD")@header)
```

## Where to go next

- `vignette("spaces-and-coordinates")` — what the affine you just read means
- `vignette("large-data")` — reading files too big to hold in memory
- `?read_vol`, `?read_header`, `?write_vol`
