---
title: Reading and writing images
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Reading and writing images}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>
---

```{r setup, include = FALSE}
source("_common.R")
```

```{r load}
library(neuroim2)
```

Every analysis starts and ends at a file. This article covers the readers and
writers, how to inspect a header before committing to loading anything, and the
one header problem that silently produces wrong results.

```{r paths}
mask_file <- system.file("extdata", "global_mask2.nii.gz", package = "neuroim2")
series_file <- system.file("extdata", "global_mask_v4.nii", package = "neuroim2")
```

## Choosing a reader

| Function | Returns | Use for |
|:--|:--|:--|
| `read_vol()` | `NeuroVol` | a 3D image |
| `read_vec()` | `NeuroVec` | a 4D series |
| `read_image()` | whichever fits | unknown dimensionality |
| `read_vol_list()` | list of `NeuroVol` | many 3D files |
| `read_header()` | `NIFTIMetaInfo` | metadata without the data |

```{r readers}
vol <- read_vol(mask_file)
vec <- read_vec(series_file)

dim(vol)
dim(vec)
```

Compression is handled by extension: `.nii` and `.nii.gz` both work, on read and
on write. `read_vol()` on a 4D file returns its first volume, so reach for
`read_image()` when you do not know what a file holds:

```{r read-image}
class(read_image(series_file))[1]
```

Several files at once, either as a list of volumes or concatenated into one
series:

```{r multi}
length(read_vol_list(c(mask_file, mask_file)))
dim(read_vec(c(series_file, series_file)))
```

## Looking before you load

`read_header()` reads metadata only. On a large file this costs nothing and
tells you what you are about to commit to:

```{r header}
hdr <- read_header(mask_file)

hdr@dims
hdr@spacing
hdr@data_type
```

The full NIfTI header sits in the `header` slot as a plain list, so anything the
accessors do not expose is still reachable:

```{r header-slot}
hdr@header$datatype
hdr@header$encoding
hdr@header$vox_offset
```

Note that geometry lives in slots on the metadata object rather than behind the
`spacing()` and `space()` generics — those are defined for images, not headers.

## The two affines

A NIfTI file can carry **two** spatial transforms: the `qform` (a rigid
transform stored as a quaternion) and the `sform` (a general affine). Each has a
code saying whether it is set. The standard says: use the sform when
`sform_code > 0`, otherwise the qform.

On a well-formed file the two agree:

```{r affines-good}
c(qform_code = hdr@header$qform_code, sform_code = hdr@header$sform_code)

all.equal(hdr@header$qform, hdr@header$sform)
```

They do not always agree, and when they disagree the file is lying to somebody.
The anatomical image shipped with this package is a real example:

```{r affines-bad}
anat_hdr <- read_header(
  system.file("extdata", "mni_downsampled.nii.gz", package = "neuroim2")
)

anat_hdr@header$pixdim[2:4]  # voxel sizes the header advertises
anat_hdr@header$qform        # 4 mm voxels
anat_hdr@header$sform        # 1 mm voxels
```

Both codes are set, so the sform wins, and the image reports 1 mm voxels even
though its `pixdim` says 4 mm:

```{r affines-consequence}
spacing(read_vol(system.file("extdata", "mni_downsampled.nii.gz", package = "neuroim2")))
```

Nothing errors. The image loads, plots correctly, and every millimetre
coordinate derived from it is wrong by a factor of four. **This is the failure
mode to know about**, because no amount of care downstream will catch it. A
two-line check when a file first arrives is worth it:

```{r affine-check}
check_affines <- function(file) {
  h <- read_header(file)
  if (h@header$qform_code > 0 && h@header$sform_code > 0) {
    isTRUE(all.equal(h@header$qform, h@header$sform, tolerance = 1e-4))
  } else {
    NA
  }
}

c(
  mask = check_affines(mask_file),
  anat = check_affines(system.file("extdata", "mni_downsampled.nii.gz", package = "neuroim2"))
)
```

## Writing

`write_vol()` and `write_vec()` take an object and a path. The extension decides
the container and the compression:

```{r write}
out_nii <- tempfile(fileext = ".nii")
out_gz <- tempfile(fileext = ".nii.gz")

write_vol(vol, out_nii)
write_vol(vol, out_gz)

c(plain = file.size(out_nii), gzipped = file.size(out_gz))
```

What matters is that a round trip preserves both the data and the geometry:

```{r roundtrip}
back <- read_vol(out_gz)

all.equal(as.array(back), as.array(vol))
all.equal(trans(space(back)), trans(space(vol)))
```

4D objects work the same way:

```{r write-vec}
out_vec <- tempfile(fileext = ".nii.gz")
write_vec(sub_vector(vec, 1:2), out_vec)

dim(read_vec(out_vec))
```

```{r cleanup, include = FALSE}
unlink(c(out_nii, out_gz, out_vec))
```

Affine coefficients are stored as 32-bit floats, so geometry survives a round
trip to about seven significant figures rather than exactly. Compare
coordinates with a tolerance, not `identical()`.

## Other formats

AFNI `HEAD`/`BRIK` pairs are read by the same functions — pass the `.HEAD` file
and the matching `.BRIK` is found alongside it. `read_meta_info()` exposes AFNI
header attributes, and NIML surface files have their own reader. `?read_vol` and
`?read_meta_info` list the supported formats.

## Where to go next

- `vignette("spaces-and-coordinates")` — what the affine you just read actually means
- `vignette("large-data")` — reading files too big to hold in memory
- `?read_vol`, `?read_header`, `?write_vol`
