---
title: Regions and searchlights
output:
  rmarkdown::html_vignette:
    css: albers.css
    includes:
      in_header: albers-header.html
vignette: |
  %\VignetteIndexEntry{Regions and searchlights}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
resource_files:
- albers.css
- albers.js
- albers-header.html
---

```{r setup, include = FALSE}
source("_common.R")
```

```{r load}
library(neuroim2)
```

Nearly every analysis in this package has the same three steps: **define a
spatial support, extract values from it, reduce them to something smaller.** What
changes is the support — one sphere, an atlas of parcels, or a sphere centred on
every voxel in turn. This article covers all three, and ends with a searchlight
analysis whose answer we can check.

```{r data}
mask <- demo_mask()
brain <- as.mask(mask > 0)
bold <- demo_bold(n_time = 40)

sum(brain)
```

## One region

`spherical_roi()` is the constructor most others build on. The centre is a voxel
index, the radius is in millimetres, and `nonzero = TRUE` discards anything
outside the mask:

```{r sphere}
roi <- spherical_roi(mask, c(32, 32, 12), radius = 8, nonzero = TRUE)

length(roi)
head(coords(roi), 3)
```

An ROI knows three things: where its voxels are (`coords()`), their linear
positions (`indices()`), and what is stored in them (`values()`). Pulling a
region's time courses out of a 4D image is `series_roi()`:

```{r series-roi}
rts <- series_roi(bold, roi)
dim(values(rts))
```

That matrix is **scans by voxels** — the transpose of `as.matrix()` on a
`NeuroVec`, so check which one you have before reducing.

Two other shapes are available. `cuboid_roi()` gives a box, `square_roi()` a
plane within one slice:

```{r other-shapes}
sp <- NeuroSpace(c(20L, 20L, 20L), c(1, 1, 1))

length(cuboid_roi(sp, c(10, 10, 10), surround = 3))
length(square_roi(sp, c(10, 10, 10), surround = 2, fixdim = 3))
```

Building many spheres at once is much faster through `spherical_roi_set()` than
a loop:

```{r roi-set}
centres <- rbind(c(20, 20, 10), c(40, 40, 14), c(32, 32, 12))
rois <- spherical_roi_set(bvol = mask, centroids = centres, radius = 8, fill = 1)

lengths(lapply(rois, indices))
```

### Combining regions

ROIs combine through their indices, so the ordinary set operations apply:

```{r set-ops}
a <- spherical_roi(mask, c(32, 32, 12), radius = 8, nonzero = TRUE)
b <- spherical_roi(mask, c(35, 32, 12), radius = 8, nonzero = TRUE)

c(
  intersection = length(intersect(indices(a), indices(b))),
  union = length(union(indices(a), indices(b))),
  only_a = length(setdiff(indices(a), indices(b)))
)
```

## Many regions: parcels

When the support comes from an atlas, a `ClusteredNeuroVol` holds the
assignment: one label per in-mask voxel.

```{r parcels}
set.seed(1)
parcels <- ClusteredNeuroVol(brain, sample(1:12, sum(brain), replace = TRUE))

num_clusters(parcels)
```

`split_clusters()` cuts a 4D image into one object per parcel:

```{r split-clusters}
parts <- split_clusters(bold, parcels)

length(parts)
dim(values(parts[[1]]))
```

`split_reduce()` goes straight to the summary, returning parcels by time in one
step — the usual way to get an atlas time-series matrix:

```{r split-reduce}
labels <- integer(prod(dim(mask)))
labels[which(as.vector(brain))] <- parcels@clusters

parcel_ts <- split_reduce(bold, factor(labels))
dim(parcel_ts)
rownames(parcel_ts)
```

`split_reduce()` wants one label per voxel in the whole grid, so out-of-mask
voxels become a group of their own — row `"0"` above, which you drop. Build the
label vector as integers and factor it at the end: assigning into an existing
factor silently produces `NA`s instead.

The default reduction is the mean; pass `FUN` for anything else.

## Every region: searchlights

A searchlight puts a sphere at every voxel in the mask. There are three
flavours, and the difference is coverage, not shape.

`searchlight()` centres one on each voxel — complete but overlapping. Built
lazily, neighbourhoods are realised only when touched:

```{r searchlight}
sl <- searchlight(brain, radius = 8, eager = FALSE, nonzero = TRUE)

length(sl)
nrow(coords(sl[[1]]))
```

`random_searchlight()` partitions the mask into *non-overlapping* spheres, which
covers every voxel exactly once for a fraction of the work:

```{r random-searchlight}
rsl <- random_searchlight(brain, radius = 8)

length(rsl)
```

`clustered_searchlight()` uses a parcellation instead of spheres, giving one
neighbourhood per parcel:

```{r clustered-searchlight}
length(clustered_searchlight(brain, cvol = parcels))
```

## A searchlight analysis that finds something

Averaging noise proves nothing, so here is a searchlight with a right answer.
We plant a task response in one sphere, then ask a searchlight to find it.

```{r plant}
design <- demo_design(40)
target <- spherical_roi(mask, c(20, 34, 12), radius = 12, nonzero = TRUE)

Y <- as.matrix(bold)
Y[indices(target), ] <- Y[indices(target), ] +
  1.3 * rep(design, each = length(target))
planted <- DenseNeuroVec(Y, space(bold))

length(target)
```

Now score each searchlight by how well its voxels track the design. This is the
map step, and it is an ordinary `sapply()` over the neighbourhoods:

```{r score}
score <- sapply(rsl, function(r) mean(cor(values(series_roi(planted, r)), design)))

range(score)
```

Writing the scores back to their voxels turns a list of numbers into an image
again:

```{r score-map}
arr <- array(0, dim(mask))
for (i in seq_along(rsl)) arr[coords(rsl[[i]])] <- score[i]
score_map <- NeuroVol(arr, space(mask))
```

Because `random_searchlight()` tiles the brain with disjoint spheres, this map is
piecewise constant — every voxel in a sphere carries its sphere's score. Use
`searchlight()` instead when you want a smoothly varying map, at the cost of one
neighbourhood per voxel rather than one per sphere.

Did it work? The peak should land inside the sphere we planted:

```{r verify}
peak <- which.max(as.vector(score_map))
peak %in% indices(target)

c(
  inside = mean(as.vector(score_map)[indices(target)]),
  outside = mean(as.vector(score_map)[setdiff(which(as.vector(brain)), indices(target))])
)
```

The peak is inside the target and mean score there is an order of magnitude
above background.

```{r score-fig, fig.cap = "Searchlight score map. The bright region is where the signal was planted.", fig.alt = "Axial slices of a searchlight correlation map with one bright cluster.", fig.height = 3.0}
plot(score_map, zlevels = c(9, 12, 15), irange = c(0, max(score)))
```

## From a map back to regions

`conn_comp()` turns a thresholded map into labelled clusters, which closes the
loop — a result becomes the support for the next analysis. Ask for
`cluster_table = TRUE` to get one row per cluster:

```{r conn-comp}
cc <- conn_comp(score_map, threshold = 0.35, cluster_table = TRUE)

head(cc$cluster_table[order(-cc$cluster_table$N), ], 4)
```

The largest cluster is an order of magnitude bigger than anything else and its
peak sits beside the centre we planted at `(20, 34, 12)`. `cc$voxels` holds the
coordinates of each one, so a cluster can become an ROI immediately.

Note that `cc$size` is a *volume*, not a per-cluster vector: every voxel holds
the size of the cluster it belongs to. Use `cluster_table$N` when you want one
number per cluster.

## Iterating without ROIs

Sometimes the pieces are the container's own: slices of a volume, volumes of a
series, or the time course at every voxel. `slices()`, `vols()` and `vectors()`
give you those directly.

```{r iterate}
anat <- demo_anatomy()

slice_means <- vapply(slices(anat), mean, numeric(1))
length(slice_means)

vol_means <- vapply(vols(bold), mean, numeric(1))
length(vol_means)
```

`vectors()` iterates voxel time courses, so mapping over it and rewrapping gives
a volume:

```{r vectors}
mean_vol <- NeuroVol(vapply(vectors(bold), mean, numeric(1)), space(mask))
dim(mean_vol)
```

## Doing it in parallel

Searchlight scoring is embarrassingly parallel. neuroim2 imports **future**, so
the change is one line of setup and `future_sapply()` in place of `sapply()`:

```{r parallel, eval = FALSE}
library(future.apply)
plan(multisession, workers = 4)

score <- future_sapply(rsl, function(r) {
  mean(cor(values(series_roi(planted, r)), design))
})

plan(sequential)
```

Keep the per-neighbourhood function self-contained — workers do not share your
session — and prefer fewer, larger tasks over many tiny ones.

## Which support should you use?

| Situation | Use |
|:--|:--|
| One hypothesis-driven region | `spherical_roi()` |
| Many known coordinates | `spherical_roi_set()` |
| A box or a single plane | `cuboid_roi()`, `square_roi()` |
| An atlas or clustering | `ClusteredNeuroVol` + `split_clusters()` |
| An atlas time-series matrix | `split_reduce()` |
| Whole-brain, overlapping | `searchlight()` |
| Whole-brain, one pass, cheaper | `random_searchlight()` |
| Whole-brain at parcel resolution | `clustered_searchlight()` |
| Clusters from a thresholded map | `conn_comp()` |

## Where to go next

- `vignette("visualization")` — drawing maps like the one above
- `vignette("large-data")` — parcel-level representations when voxelwise is too big
- `?spherical_roi`, `?searchlight`, `?split_reduce`, `?conn_comp`
