---
title: Resampling and orientation
output:
  rmarkdown::html_vignette:
    css: albers.css
    includes:
      in_header: albers-header.inc
vignette: |
  %\VignetteIndexEntry{Resampling and orientation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
resource_files:
- albers.css
- albers.js
- albers-header.inc
---

```{r setup, include = FALSE}
source("_common.R")
```

```{r load}
library(neuroim2)
```

Four operations change an image's grid. They are easy to confuse because three
of them can change its shape, but the distinction that matters is different:
whether the anatomy stays where it was in the scanner.

| Operation | Changes the array | Moves the anatomy in world space | Use when |
|:--|:--|:--|:--|
| `downsample()` | yes | no | you want fewer voxels |
| `resample_to()` | yes | no | you must match another image |
| `reorient()` | **no** | **yes** | the axis codes are wrong |
| `deoblique()` | yes | no | the affine is oblique |

The first, second and fourth rewrite voxel values precisely *in order to* leave
the anatomy where it is. `reorient()` is the opposite: it leaves the array alone
and relabels the axes, which relocates every voxel in world space. That makes it
the dangerous one if the codes were not in fact wrong.

```{r data}
anat <- demo_anatomy()
mask <- demo_mask()
```

## Downsampling

`downsample()` keeps the field of view and coarsens the grid. Give it exactly
one of `factor`, `spacing` or `outdim`:

```{r downsample}
half <- downsample(anat, factor = 0.5)

dim(anat)
dim(half)
round(spacing(half), 3)
```

The result is no longer isotropic, because `round(57 * 0.5)` is 28 rather than
28.5 — rounding a grid always perturbs the spacing slightly.

```{r downsample-fig, fig.cap = "Full and half resolution, drawn at matched physical extent.", fig.alt = "Two axial slices, the second visibly blockier.", fig.height = 3.4}
op <- par(mfrow = c(1, 2), mar = c(1, 1, 2, 1))
image(seq_len(dim(anat)[1]), seq_len(dim(anat)[2]), anat[, , 24],
  main = "original", col = gray.colors(256), axes = FALSE, ann = TRUE,
  xlab = "", ylab = "", asp = 1
)
image(seq(1, dim(anat)[1], length.out = dim(half)[1]),
  seq(1, dim(anat)[2], length.out = dim(half)[2]), half[, , 12],
  main = "factor = 0.5", col = gray.colors(256), axes = FALSE, ann = TRUE,
  xlab = "", ylab = "", asp = 1
)
par(op)
```

`downsample()` box-averages, which is right for intensity images and wrong for
labels — averaging label 3 and label 5 gives you label 4.

## Resampling onto a target grid

`resample_to()` puts an image onto any grid you name. Building that grid is
where things go wrong, so start there.

### Build the target from the source affine

It is tempting to describe a target with `dim`, `spacing` and `origin`. Do not:
those three do not carry **orientation**, and the constructor produces an
axis-aligned RAS affine regardless of what the source looked like.

```{r bad-target}
affine_to_axcodes(trans(mask))

naive <- NeuroSpace(round(dim(mask) * 1.6), spacing(mask) / 1.6, origin = origin(mask))
affine_to_axcodes(trans(naive))
```

The source runs `LAS` and the target runs `RAS`. The two affines differ only in
the sign of their first column, which is enough to put the target grid where the
source has no data. Resampling into it succeeds and returns an empty image:

```{r bad-resample}
sum(resample_to(mask, naive, method = "nearest"))
```

No error, no warning, no data. Scale the source's own affine instead, which
keeps orientation and origin intact:

```{r good-target}
tr <- trans(mask)
tr[1:3, 1:3] <- tr[1:3, 1:3] / 1.6

finer <- NeuroSpace(round(dim(mask) * 1.6), trans = tr)

affine_to_axcodes(trans(finer))
round(spacing(finer), 3)
```

Then check the result, every time. Rounding the grid means the true scale
factors are not exactly 1.6, so compare against the ratio you actually got:

```{r good-resample}
up <- resample_to(mask, finer, method = "nearest")

expected <- sum(mask) * prod(dim(finer) / dim(mask))
c(source = sum(mask), resampled = sum(up), expected = expected)

abs(sum(up) / expected - 1)
```

Under a tenth of a percent. A discrepancy much larger than that means the target
grid does not enclose the source.

### Choosing an interpolation method

`"nearest"`, `"linear"` and `"cubic"` are available and the default is
`"nearest"`, so pass `"linear"` explicitly for continuous data. The rule is
**labels and masks take `"nearest"`, continuous data takes `"linear"` or
`"cubic"`**, and the reason is one number:

```{r interpolation}
lin <- resample_to(mask, finer, method = "linear")
a <- as.array(lin)

fractional <- sum(abs(a - round(a)) > 1e-6)
c(voxels = fractional, of_resampled_mask = fractional / sum(up))
```

Nearly a fifth of the mask comes back as values strictly between 0 and 1. Those
voxels are neither in nor out, and every count taken afterwards is wrong.

Nearest is not free either: it preserves the *label set*, not the *volume*. Any
mask resample changes how many voxels are in the mask, so check `sum()` whichever
method you use.

### Matching another image

The common real task needs no target construction — pass the image whose grid
you want:

```{r match-image}
coarse <- downsample(mask, factor = 0.5) # used only as a grid donor
matched <- resample_to(mask, coarse, method = "nearest")

identical(dim(matched), dim(coarse))
identical(trans(space(matched)), trans(space(coarse)))
```

Two images that share a `NeuroSpace` can be compared, overlaid and subtracted.
Two that do not, cannot — several plotting helpers require it explicitly rather
than resampling silently behind your back.

## Reorienting

`reorient()` rewrites a space so its axes carry different codes, without
touching a single voxel. `vignette("spaces-and-coordinates")` covers the codes
themselves; what matters here is what it does to positions.

```{r reorient}
ras <- reorient(space(mask), c("R", "A", "S"))

affine_to_axcodes(trans(space(mask)))
affine_to_axcodes(trans(ras))
```

The array is untouched, so the anatomy moves:

```{r reorient-moves}
rbind(
  before = as.vector(grid_to_coord(space(mask), matrix(c(1, 1, 1), nrow = 1))),
  after = as.vector(grid_to_coord(ras, matrix(c(1, 1, 1), nrow = 1)))
)
```

Voxel `(1,1,1)` has jumped from `x = +112` to `x = -112`. Use `reorient()` when a
file's codes are genuinely wrong, not to change how an image is stored.

There is no `reorient()` method for a `NeuroVol` — it operates on spaces. To
rearrange the array as well, reorient a copy of the space and resample into it:

```{r reorient-resample}
flipped <- resample_to(mask, reorient(space(mask), c("R", "A", "S")), method = "nearest")

c(source = sum(mask), flipped = sum(flipped))
```

That is lossless here because `LAS` to `RAS` is a pure flip. It is **not**
lossless when the requested codes permute axes: `reorient()` leaves `dim` and
`spacing` alone, so a rotated box no longer encloses the original one and the
corners are cut off.

```{r reorient-permute}
permuted <- reorient(space(mask), c("P", "S", "R"))
lost <- resample_to(mask, permuted, method = "nearest")

c(source = sum(mask), permuted = sum(lost), kept = sum(lost) / sum(mask))
```

A fifth of the mask, gone without a warning. For anything beyond a flip, build a
target grid that encloses the rotated field of view — `output_aligned_space()`
and `deoblique()`'s grid arguments exist for this.

## Deobliquing

Scanners often acquire at a tilt, leaving off-diagonal terms in the affine.
`obliquity()` measures it, in radians per axis:

```{r oblique}
aff <- matrix(c(
   3.0,  0.3,  0.0,  -90,
   0.0,  3.0,  0.15, -126,
   0.0,  0.0,  4.0,  -72,
   0.0,  0.0,  0.0,    1
), nrow = 4, byrow = TRUE)

tilted <- NeuroVol(array(rnorm(64 * 64 * 30), c(64, 64, 30)),
                   NeuroSpace(c(64L, 64L, 30L), trans = aff))

round(obliquity(trans(space(tilted))) * 180 / pi, 2)
```

`deoblique()` builds an axis-aligned space enclosing the field of view and, given
a volume rather than a space, resamples the data into it.

```{r deoblique}
straight <- deoblique(tilted)

rbind(before = c(dim(tilted), round(spacing(space(tilted)), 2)),
      after = c(dim(straight), round(spacing(space(straight)), 2)))
```

The grid grows for two reasons, and only one is the tilt. An axis-aligned box
around a rotated one is bigger — but `deoblique()` also regrids to an **isotropic
grid at the smallest input voxel size** by default, which is why the third axis
went from 30 slices at 4 mm to 40 at 3 mm despite having no tilt at all. Pass a
grid explicitly if that is not what you want.

`deoblique()` also defaults to `method = "linear"`, the opposite of
`resample_to()`. On a mask, pass `method = "nearest"`:

```{r deoblique-mask}
frac <- function(x) mean(abs(as.array(x) - round(as.array(x))) > 1e-6)

c(linear = frac(deoblique(mask)), nearest = frac(deoblique(mask, method = "nearest")))
```

`as_canonical()` is the related convenience, reorienting to RAS and resampling in
one step. On an image that is already canonical it returns the input untouched,
so check the codes rather than the dimensions — `as_canonical()` never permutes
the array:

```{r canonical}
affine_to_axcodes(trans(space(mask)))
affine_to_axcodes(trans(space(as_canonical(mask))))
```

Because it is `reorient()` plus `resample()`, it inherits the enclosure problem
above. Check `sum()` afterwards on anything that is not a pure flip.

## Where to go next

- `vignette("spaces-and-coordinates")` — the affine you are rebuilding
- `vignette("visualization")` — checking a resample by eye
- `?resample_to`, `?downsample`, `?reorient`, `?deoblique`, `?as_canonical`
