---
title: Resampling and orientation
output:
  rmarkdown::html_vignette:
    css: albers.css
    includes:
      in_header: albers-header.html
vignette: |
  %\VignetteIndexEntry{Resampling and orientation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
resource_files:
- albers.css
- albers.js
- albers-header.html
---

```{r setup, include = FALSE}
source("_common.R")
```

```{r load}
library(neuroim2)
```

Four operations change an image's grid, and they are easy to confuse because
three of them can produce an image of a different shape. The distinction that
matters is whether the voxel *values* move:

| Operation | Changes the grid | Moves the data | Use when |
|:--|:--|:--|:--|
| `downsample()` | coarser, same field of view | yes | you want fewer voxels |
| `resample_to()` | any target grid | yes | you must match another image |
| `reorient()` | relabels axes only | **no** | the axis codes are wrong |
| `deoblique()` | removes scanner tilt | yes | the affine is oblique |

```{r data}
anat <- demo_anatomy()
mask <- demo_mask()
```

## Downsampling

`downsample()` keeps the field of view and coarsens the grid. Give it exactly one
of `factor`, `spacing` or `outdim`:

```{r downsample}
half <- downsample(anat, factor = 0.5)

dim(anat)
dim(half)
round(spacing(half), 3)
```

```{r downsample-fig, fig.cap = "Full resolution and half resolution, same field of view.", fig.alt = "Two axial slices, the second visibly blockier.", fig.height = 3.4}
op <- par(mfrow = c(1, 2), mar = c(1, 1, 2, 1))
image(anat[, , 24], main = "original", col = gray.colors(256), axes = FALSE, asp = 1)
image(half[, , 12], main = "factor = 0.5", col = gray.colors(256), axes = FALSE, asp = 1)
par(op)
```

The current implementation box-averages in space, which is the right default for
intensity images and the wrong one for labels — averaging label 3 and label 5
gives you label 4.

## Resampling onto a target grid

`resample_to()` puts an image onto any grid you name. Building that grid is
where things go wrong, so start there.

### Build the target from the source affine

It is tempting to describe a target with `dim`, `spacing` and `origin`. Do not:
those three do not carry **orientation**, and the constructor will give you an
axis-aligned RAS affine regardless of what the source looked like.

```{r bad-target}
affine_to_axcodes(trans(mask))

naive <- NeuroSpace(round(dim(mask) * 1.6), spacing(mask) / 1.6, origin = origin(mask))
affine_to_axcodes(trans(naive))
```

The source runs `LAS` and the target runs `RAS`, so the target grid sits
somewhere the source has no data. Resampling into it succeeds and returns an
empty image:

```{r bad-resample}
sum(resample_to(mask, naive, method = "nearest"))
```

No error, no warning, no data. Instead, scale the source's own affine, which
keeps orientation and origin intact:

```{r good-target}
tr <- trans(mask)
tr[1:3, 1:3] <- tr[1:3, 1:3] / 1.6

finer <- NeuroSpace(round(dim(mask) * 1.6), trans = tr)

affine_to_axcodes(trans(finer))
round(spacing(finer), 3)
```

```{r good-resample}
up <- resample_to(mask, finer, method = "nearest")

c(source = sum(mask), resampled = sum(up), expected = sum(mask) * 1.6^3)
```

The in-mask volume is preserved to within a percent, which is the check worth
making after any resample.

### Choosing an interpolation method

`"nearest"`, `"linear"` and `"cubic"` are available. The rule is simple: **labels
and masks take `"nearest"`, continuous data takes `"linear"` or `"cubic"`.** The
reason is visible in one number:

```{r interpolation}
non_integer <- function(x) mean(abs(as.array(x) - round(as.array(x))) > 1e-6)

c(
  nearest = non_integer(resample_to(mask, finer, method = "nearest")),
  linear = non_integer(resample_to(mask, finer, method = "linear"))
)
```

Linear interpolation turns 5% of a binary mask into fractions somewhere between
0 and 1. Those voxels are neither in nor out, and every count you take
afterwards is wrong.

### Matching another image

The common real task is putting one image onto another's grid, which needs no
target construction at all — pass the image:

```{r match-image}
coarse <- downsample(mask, factor = 0.5)
matched <- resample_to(mask, coarse, method = "nearest")

identical(dim(matched), dim(coarse))
identical(trans(space(matched)), trans(space(coarse)))
```

Two images that share a `NeuroSpace` can be compared, overlaid and subtracted.
Two that do not, cannot — several plotting helpers require it explicitly rather
than silently resampling behind your back.

## Reorienting

`reorient()` is the odd one out: it rewrites the affine so the axes carry
different codes, without touching a single voxel.

```{r reorient}
ras <- reorient(space(mask), c("R", "A", "S"))

affine_to_axcodes(trans(space(mask)))
affine_to_axcodes(trans(ras))
```

It is relative to the current orientation, so asking for the orientation an
image already has is a no-op:

```{r reorient-noop}
identical(trans(reorient(space(mask), c("L", "A", "S"))), trans(space(mask)))
```

Use it when a file's axis codes are wrong. If you want the array itself
rearranged so that, say, the first index really does run left to right, that is a
resample onto a reoriented grid — build the target with `reorient()`, then call
`resample_to()`.

## Deobliquing

Scanners often acquire at a tilt, leaving off-diagonal terms in the affine.
`obliquity()` measures it:

```{r oblique}
aff <- matrix(c(
   2.0,  0.2,  0.0,  -90,
   0.0,  2.0,  0.1, -126,
   0.0,  0.0,  2.0,  -72,
   0.0,  0.0,  0.0,    1
), nrow = 4, byrow = TRUE)

tilted <- NeuroVol(array(rnorm(40 * 40 * 30), c(40, 40, 30)),
                   NeuroSpace(c(40L, 40L, 30L), trans = aff))

round(obliquity(trans(space(tilted))), 4)
```

`deoblique()` builds an axis-aligned space enclosing the original field of view
and — when given a volume rather than a space — resamples the data into it. The
grid grows slightly, because an axis-aligned box around a tilted one is bigger:

```{r deoblique}
straight <- deoblique(tilted)

dim(tilted)
dim(straight)
round(obliquity(trans(space(straight))), 6)
```

`as_canonical()` is the related convenience for getting an image into a standard
axis order:

```{r canonical}
dim(as_canonical(anat))
```

## Where to go next

- `vignette("spaces-and-coordinates")` — what the affine you are rebuilding means
- `vignette("visualization")` — checking a resample by eye
- `?resample_to`, `?downsample`, `?reorient`, `?deoblique`
