---
title: Smoothing and filtering
output:
  rmarkdown::html_vignette:
    css: albers.css
    includes:
      in_header: albers-header.html
vignette: |
  %\VignetteIndexEntry{Smoothing and filtering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
resource_files:
- albers.css
- albers.js
- albers-header.html
---

```{r setup, include = FALSE}
source("_common.R")
```

```{r load}
library(neuroim2)
```

Filtering trades detail for noise. neuroim2 offers six filters that make that
trade differently: some blur everything, some try to keep edges, one sharpens
instead, and two work across time as well as space. This article shows what each
one does to the same image, and gives you numbers rather than adjectives.

```{r data}
anat <- read_vol(system.file("extdata", "mni_downsampled.nii.gz", package = "neuroim2"))
mask <- as.mask(anat > 100)
```

A helper to compare slices on a shared intensity scale, so differences between
panels are real rather than an artefact of rescaling:

```{r panel-helper}
compare <- function(..., z = 24) {
  imgs <- list(...)
  zlim <- range(vapply(imgs, function(v) range(as.array(v)), numeric(2)))
  op <- par(mfrow = c(1, length(imgs)), mar = c(0.5, 0.5, 2, 0.5))
  for (nm in names(imgs)) {
    image(imgs[[nm]][, , z],
      main = nm, zlim = zlim, col = gray.colors(256),
      axes = FALSE, asp = 1
    )
  }
  par(op)
}
```

## Gaussian blur

`gaussian_blur()` is the fast, unconditional baseline: every voxel becomes a
weighted average of its neighbours regardless of what is there.

```{r gaussian, fig.cap = "Gaussian smoothing at two kernel widths. Structure survives the narrow kernel and dissolves under the wide one.", fig.alt = "Three axial slices: original, lightly smoothed, heavily smoothed.", fig.height = 3.2}
light <- gaussian_blur(anat, mask, sigma = 2, window = 1)
heavy <- gaussian_blur(anat, mask, sigma = 3, window = 3)

compare(original = anat, `sigma 2, window 1` = light, `sigma 3, window 3` = heavy)
```

Two parameters, and it is worth knowing which one is doing the work. `sigma` is
the decay scale in millimetres; `window` is the discrete support in voxels, so
`window = 1` gives a 3 x 3 x 3 kernel and `window = 3` a 7 x 7 x 7 one. The
window truncates the kernel, so raising `sigma` past what the window can hold
buys you very little:

```{r sigma-window}
similarity <- function(x) cor(as.vector(as.array(x)), as.vector(as.array(anat)))

c(
  `sigma 1, window 1` = similarity(gaussian_blur(anat, mask, sigma = 1, window = 1)),
  `sigma 3, window 1` = similarity(gaussian_blur(anat, mask, sigma = 3, window = 1)),
  `sigma 3, window 3` = similarity(gaussian_blur(anat, mask, sigma = 3, window = 3))
)
```

Tripling `sigma` inside a fixed window barely moves the result; widening the
window is what actually smooths. Set both together.

## Edge-preserving filters

Gaussian blur does not know where anatomy stops. The next two do.

`guided_filter()` fits local linear models, smoothing within regions while
leaving boundaries alone. `bilateral_filter()` weights neighbours by intensity
similarity as well as distance, so voxels across a tissue boundary contribute
little.

```{r edge-preserving, fig.cap = "Gaussian blur against two edge-preserving filters at comparable strength.", fig.alt = "Four axial slices comparing original, Gaussian, guided and bilateral filtering.", fig.height = 3.2}
guided <- guided_filter(anat, radius = 4, epsilon = 0.7^2)
bilateral <- bilateral_filter(anat, spatial_sigma = 2, intensity_sigma = 1, window = 1)

compare(original = anat, gaussian = light, guided = guided, bilateral = bilateral)
```

The difference is easier to see in numbers than in a small figure. A filter that
preserves edges should reduce variance *within* tissue while keeping the overall
intensity structure intact:

```{r edge-numbers}
edge_stats <- function(x) {
  a <- as.array(x)
  c(sd = sd(a), similarity = cor(as.vector(a), as.vector(as.array(anat))))
}

round(rbind(
  gaussian = edge_stats(light),
  guided = edge_stats(guided),
  bilateral = edge_stats(bilateral)
), 4)
```

That table is the whole argument for edge-preserving filters. `guided_filter()`
removes the most variance of the three — the lowest `sd` — and yet stays
*closer* to the original image than the Gaussian does. Gaussian blur pays for
its variance reduction by moving the image; the guided filter gets more
reduction for less distortion, because it spends its smoothing inside regions
instead of across their boundaries. `bilateral_filter()` is the most
conservative here on both counts.

Tune `epsilon` and `intensity_sigma` to move along that trade: smaller values
preserve more contrast, larger values converge on plain Gaussian behaviour.

## Sharpening

`laplace_enhance()` runs the other direction, amplifying detail with a
multi-scale non-local scheme. It is a display and detection aid, not a
denoiser — it raises noise along with structure.

```{r sharpen, fig.cap = "Laplacian enhancement increases local contrast rather than reducing it.", fig.alt = "Two axial slices: original and sharpened.", fig.height = 3.4}
sharp <- laplace_enhance(anat, k = 2, patch_size = 3, search_radius = 1, h = 0.7)

compare(original = anat, enhanced = sharp)
```

```{r sharpen-numbers}
c(original = sd(as.array(anat)), enhanced = sd(as.array(sharp)))
```

Variance goes up, which is the point. Raise `k` or `h` for a stronger effect,
and watch the background — that is where over-enhancement shows first.

## Filtering across time

Two filters work on a `NeuroVec`, smoothing over space and time together. Both
need a series with real temporal structure, so we simulate one:

```{r bold}
bold_mask <- read_vol(system.file("extdata", "global_mask2.nii.gz", package = "neuroim2"))
bold <- simulate_fmri(bold_mask, n_time = 40, seed = 1)
```

`bilateral_filter_4d()` extends the bilateral idea along the time axis, with
`temporal_sigma` and `temporal_window` controlling that extent and
`temporal_spacing` giving the units of the time axis (the TR, if you are working
in seconds).

```{r bilat4d}
bf4d <- bilateral_filter_4d(
  bold, bold_mask,
  spatial_sigma = 2, intensity_sigma = 1, temporal_sigma = 1,
  spatial_window = 1, temporal_window = 1, temporal_spacing = 1
)
```

`cgb_filter()` decides which voxels belong together by **correlating their time
courses**, not by comparing intensities. It builds a sparse graph — spatial
proximity times time-series affinity — and diffuses the data over it. That makes
it the right choice for functional data, where two voxels in the same network
should smooth together even if their raw intensities differ.

```{r cgb}
cgb <- cgb_filter(
  bold,
  mask = bold_mask,
  spatial_sigma = 5, window = NULL,
  corr_map = "power", corr_param = 2,
  topk = 16, passes = 1, lambda = 1
)
```

Both should reduce per-voxel noise and raise the coherence of nearby voxels.
They do:

```{r fourd-numbers}
inmask <- head(which(as.vector(bold_mask) > 0), 300)

summarise <- function(x) {
  m <- as.matrix(x)[inmask, ]
  c(
    voxel_sd = mean(apply(m, 1, sd)),
    pairwise_cor = mean(cor(t(m))[upper.tri(diag(nrow(m)))])
  )
}

round(rbind(
  raw = summarise(bold),
  bilateral_4d = summarise(bf4d),
  cgb = summarise(cgb)
), 3)
```

Per-voxel standard deviation falls and mean pairwise correlation rises, with
`cgb_filter()` doing more of both at these settings. Raise `passes`, or lower
`lambda`, to move further along that trade.

`return_graph = TRUE` keeps the graph so you can reuse it instead of rebuilding
it for every variant:

```{r cgb-graph}
out <- cgb_filter(bold, bold_mask, spatial_sigma = 5, topk = 16, return_graph = TRUE)

names(out)
```

## Choosing

| Want | Use | Watch out for |
|:--|:--|:--|
| A fast, predictable baseline | `gaussian_blur()` | set `window` as well as `sigma` |
| Denoise but keep tissue boundaries | `guided_filter()` | `epsilon` sets the trade |
| Same, with finer intensity control | `bilateral_filter()` | cost grows with `window` |
| More visible detail, not less | `laplace_enhance()` | amplifies noise too |
| Joint space and time denoising | `bilateral_filter_4d()` | `temporal_spacing` must match your TR |
| Smoothing that follows networks | `cgb_filter()` | needs series with real temporal structure |

## Where to go next

- `vignette("visualization")` — inspecting the results of any of these
- `vignette("resampling-and-orientation")` — changing the grid rather than the values
- `?gaussian_blur`, `?guided_filter`, `?cgb_filter`, `?cgb_make_graph`
