---
title: Spaces and coordinates
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Spaces and coordinates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>
---

```{r setup, include = FALSE}
source("_common.R")
```

```{r load}
library(neuroim2)
```

Every object in neuroim2 — a `NeuroVol`, a `NeuroVec`, an ROI — carries a
`NeuroSpace` that answers one question: *where in the scanner does this voxel
sit?* Get that mapping right and anatomical coordinates, atlas overlays and
multi-subject registration all follow. Get it wrong and your activations end up
in the other hemisphere.

## What a NeuroSpace holds

Call `space()` on any object to see it. Here is the space of the brain mask that
ships with the package:

```{r real-space}
mask <- demo_mask()
space(mask)
```

Five things are recorded:

- **Grid dimensions** — voxel counts along each axis (`dim`)
- **Voxel spacing** — physical voxel size in millimetres (`spacing`)
- **Origin** — world coordinates of voxel `(1, 1, 1)` (`origin`)
- **Affine transform** — the 4 x 4 matrix mapping voxel indices to millimetres (`trans`)
- **Axis orientation** — the anatomical direction each axis runs (`axes`)

You can also build one directly, which is what the rest of this article does so
that the numbers stay easy to follow:

```{r create-space}
sp <- NeuroSpace(
  dim     = c(64L, 64L, 40L),
  spacing = c(2, 2, 2),
  origin  = c(-90, -126, -72)
)

dim(sp)
spacing(sp)
origin(sp)
```

## The affine transform

The affine is a 4 x 4 homogeneous matrix in `trans(sp)`. It maps a
**zero-based** voxel position to millimetres:

```
[x_mm]   [        ] [i]
[y_mm] = [  M   t ] [j]
[z_mm]   [        ] [k]
[  1 ]   [ 0    1 ] [1]
```

`M` is the 3 x 3 linear block — spacing, rotation and any shear — and `t` is the
translation, which is the origin. For an axis-aligned image `M` is diagonal:

```{r show-trans}
trans(sp)
```

Both pieces can be read straight back out:

```{r decompose-affine}
trans(sp)[1:3, 4]          # translation column = origin
diag(trans(sp)[1:3, 1:3])  # diagonal of linear block = voxel sizes
```

The inverse (world to voxel) is cached on the space:

```{r inverse-trans}
inverse_trans(sp)
```

You can supply a full affine instead of `spacing` and `origin`, and neuroim2
derives the rest from it:

```{r explicit-affine}
aff <- diag(c(3, 3, 4, 1))
aff[1:3, 4] <- c(-90, -126, -72)

sp_aff <- NeuroSpace(dim = c(60L, 60L, 35L), trans = aff)
spacing(sp_aff)
origin(sp_aff)
```

## Three ways to address a voxel

neuroim2 uses **1-based grid indices** throughout, matching R's arrays. There
are two voxel addressing schemes and one world system:

| Address | Range | Description |
|:--|:--|:--|
| Linear index | `1 ... prod(dim)` | one integer, column-major, as in any R array |
| Grid index | `(1...d1, 1...d2, 1...d3)` | a 1-based triple |
| World coordinate | millimetres | defined by the affine |

```{r coord-diagram, echo = FALSE, fig.cap = "The three addressing schemes and the functions that convert between them.", fig.alt = "Diagram showing linear index, grid index and world coordinates connected by conversion functions.", fig.width = 6.4, fig.height = 2.4}
op <- par(mar = c(0, 0, 0, 0))
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0.4, 2.6))

rect(0.2, 0.8, 2.8, 2.2, col = "#dce8f5", border = "#3a7abf", lwd = 1.5)
rect(3.7, 0.8, 6.3, 2.2, col = "#dce8f5", border = "#3a7abf", lwd = 1.5)
rect(7.2, 0.8, 9.8, 2.2, col = "#dce8f5", border = "#3a7abf", lwd = 1.5)

text(1.5, 1.7, "Linear\nindex", cex = 0.85, font = 2)
text(1.5, 1.15, "1 ... prod(dim)", cex = 0.68, col = "#555555")
text(5.0, 1.7, "Grid\nindex", cex = 0.85, font = 2)
text(5.0, 1.15, "(i, j, k)  1-based", cex = 0.68, col = "#555555")
text(8.5, 1.7, "World\ncoords", cex = 0.85, font = 2)
text(8.5, 1.15, "x, y, z  mm", cex = 0.68, col = "#555555")

arrows(2.85, 1.5, 3.65, 1.5, length = 0.08, lwd = 1.4, col = "#3a7abf")
arrows(3.65, 1.3, 2.85, 1.3, length = 0.08, lwd = 1.4, col = "#888888")
text(3.25, 1.78, "index_to_grid", cex = 0.58, col = "#3a7abf")
text(3.25, 1.02, "grid_to_index", cex = 0.58, col = "#888888")

arrows(6.35, 1.5, 7.15, 1.5, length = 0.08, lwd = 1.4, col = "#3a7abf")
arrows(7.15, 1.3, 6.35, 1.3, length = 0.08, lwd = 1.4, col = "#888888")
text(6.75, 1.78, "grid_to_coord", cex = 0.58, col = "#3a7abf")
text(6.75, 1.02, "coord_to_grid", cex = 0.58, col = "#888888")

arrows(2.85, 0.72, 7.15, 0.72, length = 0.08, lwd = 1.2, col = "#3a7abf", lty = 2)
arrows(7.15, 0.52, 2.85, 0.52, length = 0.08, lwd = 1.2, col = "#888888", lty = 2)
text(5.0, 0.85, "index_to_coord", cex = 0.55, col = "#3a7abf")
text(5.0, 0.42, "coord_to_index", cex = 0.55, col = "#888888")

par(op)
```

### Grid and linear index

```{r grid-index}
grid_to_index(sp, matrix(c(10, 12, 5), nrow = 1))
index_to_grid(sp, 17098L)
```

### Grid and world

`grid_to_coord()` subtracts 1 from the 1-based grid before applying the affine,
so voxel `(1, 1, 1)` lands exactly on the origin:

```{r grid-to-coord}
grid_to_coord(sp, matrix(c(1, 1, 1), nrow = 1))
origin(sp)
```

Pass a matrix with one row per point to convert many at once:

```{r grid-to-coord-multi}
pts <- matrix(c(
   1,  1,  1,
  32, 32, 20,
  64, 64, 40
), ncol = 3, byrow = TRUE)

grid_to_coord(sp, pts)
```

### World back to grid

```{r coord-to-grid}
coord_to_grid(sp, c(0, 0, 0))
```

### A round trip

Going out to millimetres and back returns the voxel you started from:

```{r roundtrip}
idx   <- 12345L
grid  <- index_to_grid(sp, idx)
world <- grid_to_coord(sp, grid)
back  <- grid_to_index(sp, coord_to_grid(sp, world))

c(index = idx, back = back)
```

## Orientation codes

Images are stored in many orientations. The orientation code names the
anatomical direction each axis runs *towards*:

| Letter | Increasing index moves towards |
|:--:|:--|
| R / L | Right / Left |
| A / P | Anterior / Posterior |
| S / I | Superior / Inferior |

`"RAS"` means axis 1 runs towards the right, axis 2 towards the front, axis 3
towards the top — the NIfTI and MNI convention. Read it straight from an affine:

```{r axcodes}
affine_to_axcodes(trans(sp))
affine_to_axcodes(trans(mask))
```

The mask is `LAS`: its first axis runs the other way, towards the left. That
single difference is the usual reason an overlay comes out mirrored, and it is
worth checking with `affine_to_axcodes()` before trusting any two images to line
up.

Changing orientation, rather than just reading it, is a separate operation —
see `vignette("resampling-and-orientation")`.

## Oblique affines

Scanners often acquire at a slight tilt, which puts off-diagonal terms in the
affine:

```{r oblique}
aff_obl <- matrix(c(
   2.0,  0.2,  0.0,  -90,
   0.0,  2.0,  0.1, -126,
   0.0,  0.0,  2.0,  -72,
   0.0,  0.0,  0.0,    1
), nrow = 4, byrow = TRUE)

sp_obl <- NeuroSpace(dim = c(91L, 109L, 91L), trans = aff_obl)
```

`spacing()` returns the **column norms** of the linear block — the true physical
edge lengths — which is why it disagrees with the diagonal here:

```{r oblique-spacing}
spacing(sp_obl)
diag(aff_obl[1:3, 1:3])
```

Always use `spacing()`, never the diagonal. `obliquity()` quantifies the tilt,
and `voxel_sizes()` computes edge lengths from any affine:

```{r obliquity}
obliquity(aff_obl)
obliquity(trans(sp))
voxel_sizes(aff_obl)
```

## Adding and dropping a time axis

`add_dim()` extends a 3D space to 4D; `drop_dim()` reverses it. Both leave the
spatial affine untouched, which is what makes it safe to move between a series
and a summary volume:

```{r dims}
sp_4d <- add_dim(sp, 200L)
dim(sp_4d)

sp_back <- drop_dim(sp_4d)
identical(trans(sp_back), trans(sp))
```

## Four things that catch people out

**Oblique affines.** If `diag(trans(sp)[1:3, 1:3])` disagrees with `spacing(sp)`,
the image is tilted. Use `obliquity()` to measure it and `deoblique()` to remove
it.

**1-based indices.** `grid_to_coord()` subtracts 1 for you. If you ever do raw
affine arithmetic yourself, subtract it first or every coordinate is off by one
voxel.

**sform versus qform.** NIfTI stores two affines. neuroim2 follows the standard
priority — use the sform when `sform_code > 0`, otherwise the qform — so
`trans(space(img))` reflects whichever won. When a file's two affines disagree,
what you get may not be what the pixdim field advertises;
`vignette("reading-and-writing")` shows how to inspect both.

**Float32 precision.** NIfTI stores affine coefficients as 32-bit floats, and
neuroim2 rounds to 7 significant figures to match. Round trips can show
floating-point noise around 0.001 mm. Compare coordinates with a tolerance.

## Quick reference

| Function | Maps | Typical use |
|:--|:--|:--|
| `grid_to_index(sp, m)` | grid to linear | looking up voxel data |
| `index_to_grid(sp, i)` | linear to grid | interpreting array subscripts |
| `grid_to_coord(sp, m)` | grid to mm | reporting a location |
| `coord_to_grid(sp, m)` | mm to grid | atlas and seed lookup |
| `affine_to_axcodes(a)` | affine to `"RAS"` | orientation check |
| `voxel_sizes(a)` | affine to mm | physical voxel size |
| `obliquity(a)` | affine to radians | tilt check |
| `add_dim(sp, n)` | 3D to 4D | attach a time axis |
| `drop_dim(sp)` | 4D to 3D | strip a time axis |

## Where to go next

- `vignette("volumes-and-vectors")` — the containers this geometry is attached to
- `vignette("resampling-and-orientation")` — changing a grid rather than just
  reinterpreting it
- `?NeuroSpace`, `?affine_to_axcodes`, `?obliquity`
