---
title: Visualization
output:
  rmarkdown::html_vignette:
    css: albers.css
    includes:
      in_header: albers-header.html
vignette: |
  %\VignetteIndexEntry{Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
resource_files:
- albers.css
- albers.js
- albers-header.html
---

```{r setup, include = FALSE}
source("_common.R")
knitr::opts_chunk$set(fig.width = 8, dpi = 120)
```

```{r load}
library(neuroim2)
```

Before anything else, you need to answer one question: *does this look right?*
The plotting helpers exist for that first visual check. They take `NeuroVol`
objects, preserve aspect ratio, scale intensities robustly by default, and
return ordinary `ggplot2` objects you can modify or save.

```{r data}
anat <- demo_anatomy()
stat_map <- demo_stat_map(demo_anatomy_mask())

identical(space(anat), space(stat_map))
```

The statistical map here is fitted, not drawn: a design is planted in two
spheres of a simulated series and a t-statistic is computed at every voxel. It
shares a `NeuroSpace` with the anatomy, which every overlay helper requires.

```{r zlevels}
zlevels <- round(seq(12, 38, length.out = 6))
```

## A montage of slices

`plot_montage()` scans several slices at once and returns a single `ggplot`
object.

```{r montage, fig.cap = "Six axial slices with robust intensity scaling.", fig.alt = "Six axial anatomical slices on a dark background.", fig.height = 3.4, dev.args = list(bg = "#141414")}
plot_montage(anat,
  zlevels = zlevels, ncol = 6,
  cmap = anatomy_cmap, range = "robust",
  title = "Anatomical montage", style = "dark"
)
```

`style = "light"` is the version for a manuscript page:

```{r montage-light, fig.cap = "The same slices in the light style.", fig.alt = "Four axial anatomical slices on a light background.", fig.height = 3.4}
plot_montage(anat,
  zlevels = zlevels[1:4], ncol = 4,
  cmap = anatomy_cmap, range = "robust",
  title = "Light style", style = "light"
)
```

`range = "robust"` clips to percentiles rather than the extremes. Use it unless
the outliers are the point — a single bright voxel on the default full range
flattens everything else to black.

## One location in three planes

`plot_ortho()` is for when you care about a specific point rather than a set of
slices. The crosshairs confirm that all three panels show the same place.

```{r ortho, fig.cap = "Axial, coronal and sagittal views through one voxel.", fig.alt = "Three orthogonal slices with crosshairs.", fig.height = 3.6, dev.args = list(bg = "#141414")}
plot_ortho(anat,
  coord = round(dim(anat) / 2), unit = "index",
  cmap = anatomy_cmap, title = "Three-plane view", style = "dark"
)
```

`draw = FALSE` returns the panels instead of drawing them, which is what you
want when composing or saving figures yourself:

```{r ortho-panels}
names(plot_ortho(anat, coord = round(dim(anat) / 2), draw = FALSE))
```

## Overlaying a statistical map

`plot_overlay()` puts one volume on top of another. For a signed map, a
diverging palette with symmetric limits keeps positive and negative comparable:

```{r overlay, fig.cap = "Thresholded signed t-map over the anatomy.", fig.alt = "Six anatomical slices with blue and orange statistical clusters.", fig.height = 3.6, dev.args = list(bg = "#141414")}
plot_overlay(
  bgvol = anat, overlay = stat_map,
  zlevels = zlevels, ncol = 6,
  bg_cmap = anatomy_cmap, bg_range = "robust",
  ov_cmap = "coldhot", ov_thresh = 3, ov_symmetric = TRUE, ov_alpha = 0.8,
  title = "t > 3", style = "dark"
)
```

The threshold is doing real work. It is worth knowing how much survives it:

```{r overlay-counts}
a <- as.array(stat_map)
c(positive = sum(a > 3), negative = sum(a < -3), total_brain = sum(a != 0))
```

`ov_symmetric = TRUE` forces the colour scale to be centred on zero. Without it,
an asymmetric map assigns different colours to `+4` and `-4`, which is
misleading with a diverging palette.

### Cleaning up a noisy map

Unsmoothed statistical maps often look like salt and pepper. `enhance_stat_map()`
despikes, smooths in an edge-preserving way, and then sharpens what survives, so
clusters read clearly at figure size:

```{r enhance}
enhanced <- enhance_stat_map(stat_map)

raw <- as.array(stat_map)
enh <- as.array(enhanced)

round(rbind(
  raw = c(peak = max(raw), above_3 = sum(raw > 3), null_sd = sd(raw[raw != 0 & abs(raw) < 2])),
  enhanced = c(max(enh), sum(enh > 3), sd(enh[enh != 0 & abs(enh) < 2]))
), 3)
```

Read that table carefully, because it is the reason this is a display tool.
The peak rises by a third and a third more voxels clear `t > 3`, while the noise
floor's spread is unchanged. Clusters are being amplified, not revealed. Thresholds
and voxel counts taken from an enhanced map mean nothing — enhance for the figure,
infer from the original.

`plot_overlay()` and `plot_ortho()` take `enhance = TRUE` to apply it inline.

## Checking a registration

Two images on the same grid can be compared directly. `plot_checkerboard()`
alternates tiles between them, so misalignment shows up as anatomy breaking
across tile edges.

```{r checkerboard, fig.cap = "Checkerboard against a copy shifted by four voxels. Anatomy breaks at tile boundaries.", fig.alt = "Checkerboard slices with visible discontinuities.", fig.height = 3.6, dev.args = list(bg = "#141414")}
shifted <- demo_shifted(anat, by = 4L)

plot_checkerboard(anat, shifted,
  zlevels = zlevels, tile = 8, ncol = 6,
  cmap = anatomy_cmap, title = "Checkerboard QC", style = "dark"
)
```

`plot_edge_overlay()` compares boundaries instead, which is easier to judge when
the two images have different contrast:

```{r edges, fig.cap = "Fixed and moving edge maps. Where the colours separate, the images disagree.", fig.alt = "Edge overlay slices with two colours of outline separating.", fig.height = 3.6, dev.args = list(bg = "#141414")}
brain <- demo_anatomy_mask()
fixed_edges <- demo_edges(brain)
moving_edges <- demo_edges(demo_shifted(brain, by = 4L))

plot_edge_overlay(anat, fixed_edges, moving_edges,
  zlevels = zlevels, ncol = 6,
  bg_cmap = anatomy_cmap, title = "Edge overlay QC", style = "dark"
)
```

Both helpers **require** matching `NeuroSpace` objects and error otherwise. That
is deliberate: a QC plot that silently resampled would hide the very problem you
are looking for. Resample first — see
`vignette("resampling-and-orientation")`.

## Colours

`resolve_cmap()` turns a palette name into colours, and `mapToColors()` maps
data onto one. Together they cover the cases the helpers do not:

```{r colours}
coldhot <- resolve_cmap("coldhot")
head(coldhot, 3)

strongest <- head(order(-abs(raw)), 5)
raw[strongest]
mapToColors(raw[strongest], col = coldhot)
```

Positive peaks come back warm, negative ones cold. Values at or below the
`threshold` argument map to `zero_col`, which defaults to fully transparent —
that is how overlays leave the background showing through.

A palette is a plain character vector, so any vector of colours works: the
`anatomy_cmap` used throughout this article is three greys.

## Which helper?

| Task | Helper |
|:--|:--|
| Quick look at one volume | `plot()` |
| Publication slice grid | `plot_montage()` |
| One location in three planes | `plot_ortho()` |
| Statistic or mask over anatomy | `plot_overlay()` |
| Registration, two images | `plot_checkerboard()` |
| Registration, two edge maps | `plot_edge_overlay()` |
| Tidy a noisy map for display | `enhance_stat_map()` |
| Consistent ggplot styling | `theme_neuro()` |

Four habits worth keeping: `range = "robust"` for anatomy, `ov_symmetric = TRUE`
for signed maps, `style = "dark"` to inspect and `"light"` to publish, and
`draw = FALSE` whenever you want to arrange panels yourself.

## Where to go next

- `vignette("regions-and-searchlights")` — producing maps worth plotting
- `vignette("resampling-and-orientation")` — getting two images onto one grid
- `?plot_montage`, `?plot_overlay`, `?enhance_stat_map`, `?theme_neuro`
