---
title: Volumes and vectors
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Volumes and vectors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-red');});</script>
---

```{r setup, include = FALSE}
source("_common.R")
```

```{r load}
library(neuroim2)
```

Two containers carry almost everything in neuroim2: `NeuroVol` for one 3D image
and `NeuroVec` for a stack of them sharing a spatial frame. This article covers
how to build them, take them apart, and put them back together.

```{r data}
anat <- read_vol(system.file("extdata", "mni_downsampled.nii.gz", package = "neuroim2"))
mask <- read_vol(system.file("extdata", "global_mask2.nii.gz", package = "neuroim2"))
bold <- simulate_fmri(mask, n_time = 20, seed = 1)

class(anat)[1]
class(bold)[1]
```

`read_vol()` reads a 3D file; `read_vec()` reads a 4D one. Here the functional
series is simulated rather than read, so that its time courses have realistic
temporal structure.

## They behave like arrays

Indexing, arithmetic and comparison all work, and — this is the important part —
the result keeps its geometry instead of decaying into a bare array:

```{r arraylike}
dim(anat)
anat[24, 28, 24]

class(anat * 2)[1]
max(anat)
```

`Arith`, `Compare`, `Logic` and `Summary` group generics are all defined, so
`anat / 2`, `anat > 100` and `range(anat)` do what you expect.

## Masks

A comparison produces a `LogicalNeuroVol` — a mask that knows where it is:

```{r mask-class}
brain <- anat > 100
class(brain)[1]
sum(brain)
```

`as.mask()` is the explicit constructor, and takes either a logical volume or a
set of indices:

```{r as-mask}
m1 <- as.mask(anat > 100)
m2 <- as.mask(anat, which(anat > 500))

c(loose = sum(m1), strict = sum(m2))
```

Masks index volumes directly, which is the usual way to pull out the voxels you
care about:

```{r mask-index}
vals <- anat[brain]

length(vals)
mean(vals)
```

```{r mask-fig, fig.cap = "A volume and the mask derived from it, same grid, same geometry.", fig.alt = "Two axial slices side by side: an anatomical slice and its binary mask.", fig.height = 3.4}
op <- par(mfrow = c(1, 2), mar = c(1, 1, 2, 1))
image(anat[, , 24], main = "anat", col = gray.colors(256), axes = FALSE, asp = 1)
image(brain[, , 24], main = "anat > 100", col = gray.colors(2), axes = FALSE, asp = 1)
par(op)
```

## Building one by hand

A `NeuroVol` is an array plus a `NeuroSpace`:

```{r build-vol}
sp <- NeuroSpace(dim = c(16L, 16L, 8L), spacing = c(2, 2, 2))
vol <- NeuroVol(array(rnorm(16 * 16 * 8), c(16, 16, 8)), sp)

vol
```

In practice you rarely write the space out. Copy it from an image you already
have, which is what keeps a derived result aligned with its source:

```{r copy-space}
derived <- NeuroVol((anat + 1) * 25, space(anat))
identical(space(derived), space(anat))
```

The same for 4D, from either an array or a voxels-by-time matrix:

```{r build-vec}
sp4 <- NeuroSpace(c(16L, 16L, 8L, 5L), spacing = c(2, 2, 2))

v_arr <- NeuroVec(array(rnorm(16 * 16 * 8 * 5), c(16, 16, 8, 5)), sp4)
v_mat <- DenseNeuroVec(matrix(rnorm(16 * 16 * 8 * 5), nrow = 16 * 16 * 8), sp4)

identical(dim(v_arr), dim(v_mat))
dim(v_arr)
```

## Taking a 4D object apart

Three ways, depending on what you want back.

A single volume, by position:

```{r extract-vol}
dim(bold[[3]])
class(bold[[3]])[1]
```

A shorter series, keeping it 4D:

```{r sub-vector}
dim(sub_vector(bold, 1:5))
```

Every volume as a list, for iteration:

```{r vols}
length(vols(bold))
```

`slices()` does the same along the third axis of a 3D image, and `slice()` pulls
one out as a `NeuroSlice`:

```{r slices}
length(slices(anat))
class(slice(anat, 24, along = 3))[1]
```

## The matrix view

`as.matrix()` flattens a `NeuroVec` to **voxels by time**, which is the shape
most modelling code wants:

```{r as-matrix}
mat <- as.matrix(bold)
dim(mat)
```

Going back needs the matching 3D space, which `drop_dim()` supplies. This
round trip — flatten, reduce along time, rewrap — is the single most common
operation in the package:

```{r reduce}
ar1 <- NeuroVol(
  apply(mat, 1, function(x) if (sd(x) > 0) cor(x[-1], x[-length(x)]) else 0),
  drop_dim(space(bold))
)

dim(ar1)
range(ar1)
```

That is a lag-1 autocorrelation map: one number per voxel, summarising how
strongly each time course predicts itself one scan later. Any function from a
numeric vector to a scalar works in its place — `mean`, `sd`, a trimmed mean,
the slope of a regression.

## Putting objects together

`concat()` stacks along time. Volumes become a series:

```{r concat-vols}
dim(concat(anat, anat, anat))
```

and series extend each other, which is how runs get joined:

```{r concat-vecs}
run1 <- sub_vector(bold, 1:5)
run2 <- sub_vector(bold, 6:12)

dim(concat(run1, run2))
```

All inputs must share a `NeuroSpace`; `concat()` errors rather than guessing if
they do not.

`split_blocks()` is the inverse, cutting a concatenated series back into runs
given a block label per timepoint:

```{r split-blocks}
blocks <- split_blocks(bold, rep(1:4, each = 5))

length(blocks)
dim(blocks[[1]])
```

## Time series at a location

`series()` takes a voxel, `series_roi()` takes a region:

```{r series}
length(series(bold, 32, 32, 12))
```

Regions, parcels and searchlights are the subject of
`vignette("regions-and-searchlights")`; this is just the hook they hang on.

## Dense and sparse

When a mask defines which voxels count, a sparse representation stores only
those:

```{r sparse}
sparse <- as.sparse(bold, as.mask(mask > 0))

class(sparse)[1]
dim(sparse)
```

Dimensions are unchanged — sparsity is about storage, not shape — and
`as.dense()` reverses it. `read_vec(file, mask = ...)` reads straight into the
sparse form without materialising the dense one first.

```{r dense-again}
class(as.dense(sparse))[1]
```

When that trade is worth making, and the other backends available when data
outgrows memory, is the subject of `vignette("large-data")`.

## Where to go next

- `vignette("reading-and-writing")` — getting these objects to and from disk
- `vignette("regions-and-searchlights")` — ROIs, parcels, searchlights
- `vignette("large-data")` — sparse, mapped and file-backed storage

Reference: `?NeuroVol`, `?NeuroVec`, `?concat`, `?sub_vector`, `?as.mask`.
