---
title: Volumes and vectors
output:
  rmarkdown::html_vignette:
    css: albers.css
    includes:
      in_header: albers-header.inc
vignette: |
  %\VignetteIndexEntry{Volumes and vectors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
resource_files:
- albers.css
- albers.js
- albers-header.inc
---

```{r setup, include = FALSE}
source("_common.R")
```

```{r load}
library(neuroim2)
```

Two containers carry almost everything in neuroim2: `NeuroVol` for one 3D image
and `NeuroVec` for a stack of them sharing a spatial frame. This article covers
how to build them, take them apart, and put them back together.

```{r data}
anat <- demo_anatomy()
mask <- demo_mask()
bold <- demo_bold(n_time = 20)

class(anat)[1]
class(bold)[1]
```

`demo_anatomy()` and friends are shorthand this article set defines for the
shipped example data; behind them are `read_vol()` for a 3D file, `read_vec()`
for a 4D one, and `simulate_fmri()` for the functional series, which is
simulated rather than read so its time courses have realistic temporal
structure.

Note the classes: `NeuroVol` and `NeuroVec` are the generic constructors, and
what you get back is a concrete implementation — `DenseNeuroVol` here, with the
sparse and on-disk alternatives covered in `vignette("large-data")`.

## What survives an operation

These objects behave like arrays, but not every operation returns one that still
knows where it is. The rule is worth learning early, because it decides when you
have to reattach a space by hand.

Arithmetic and comparison preserve geometry — `Arith`, `Compare` and `Logic`
group generics are all defined:

```{r arraylike}
class(anat * 2)[1]
class(anat > 100)[1]
```

Extraction deliberately does not. `[` gives you plain R data:

```{r indexing}
dim(anat)
anat[24, 28, 24]

class(anat[, , 24])[1]
class(anat[anat > 100])[1]
```

That is the right default — you usually want a matrix or a vector to compute on
— but it means anything you build from extracted values has to be given a space
again, which is the subject of the next section but one.

## Masks

A comparison produces a `LogicalNeuroVol` — a mask that knows where it is:

```{r mask-class}
brain <- anat > 100
class(brain)[1]
sum(brain)
```

`as.mask()` is the explicit constructor, and takes either a logical volume or a
set of indices:

```{r as-mask}
from_logical <- as.mask(anat > 100)
from_indices <- as.mask(anat, which(anat > 6000))

c(brain = sum(from_logical), bright = sum(from_indices))
```

Masks index volumes directly, which is the usual way to pull out the voxels you
care about:

```{r mask-index}
vals <- anat[brain]

length(vals)
mean(vals)
```

```{r mask-fig, fig.cap = "A volume and the mask derived from it, same grid, same geometry.", fig.alt = "Two axial slices side by side: an anatomical slice and its binary mask.", fig.height = 3.4}
op <- par(mfrow = c(1, 2), mar = c(1, 1, 2, 1))
image(anat[, , 24], main = "anat", col = gray.colors(256), axes = FALSE, asp = 1)
image(brain[, , 24], main = "anat > 100", col = gray.colors(2), axes = FALSE, asp = 1)
par(op)
```

## Building one by hand

A `NeuroVol` is an array plus a `NeuroSpace`:

```{r build-vol}
set.seed(1)

sp <- NeuroSpace(dim = c(16L, 16L, 8L), spacing = c(2, 2, 2))
vol <- NeuroVol(array(rnorm(16 * 16 * 8), c(16, 16, 8)), sp)

vol
```

In practice you rarely write a space out. Arithmetic keeps the one it has, so
the case that matters is rebuilding an image from values that have *lost* their
geometry — anything that has been through `[`, `as.matrix()` or `apply()`:

```{r copy-space}
values <- as.array(anat)[]        # a plain numeric vector: no space
class(values)

derived <- NeuroVol(values, space(anat))
identical(space(derived), space(anat))
```

The same for 4D, from either an array or a voxels-by-time matrix:

```{r build-vec}
sp4 <- NeuroSpace(c(16L, 16L, 8L, 5L), spacing = c(2, 2, 2))
d <- rnorm(16 * 16 * 8 * 5)

v_arr <- NeuroVec(array(d, c(16, 16, 8, 5)), sp4)
v_mat <- NeuroVec(matrix(d, nrow = 16 * 16 * 8), sp4)

dim(v_arr)
all.equal(as.array(v_arr), as.array(v_mat))
```

## Taking an object apart

A few ways, depending on what you want back.

A single volume, by position:

```{r extract-vol}
dim(bold[[3]])
class(bold[[3]])[1]
```

A shorter series, keeping it 4D:

```{r sub-vector}
dim(sub_vector(bold, 1:5))
```

Every volume as a list, for iteration:

```{r vols}
length(vols(bold))
```

For a 3D image the equivalent is `slices()` along the third axis, with `slice()`
pulling one out as a `NeuroSlice`:

```{r slices}
length(slices(anat))
class(slice(anat, 24, along = 3))[1]
```

## The matrix view

`as.matrix()` flattens a `NeuroVec` to **voxels by time**, which is the shape
most modelling code wants:

```{r as-matrix}
mat <- as.matrix(bold)
dim(mat)
```

Most of those rows are outside the brain and constant, so reduce over the mask
and scatter the answers back. This masked form is the pattern you will use most
often, and it is what the rest of these articles assume:

```{r reduce}
inside <- which(as.vector(mask) > 0)

ar1 <- numeric(nrow(mat))
ar1[inside] <- apply(mat[inside, ], 1, function(x) cor(x[-1], x[-length(x)]))

ar1_map <- NeuroVol(ar1, drop_dim(space(bold)))

c(voxels = nrow(mat), reduced = length(inside))
round(range(ar1[inside]), 3)
```

`drop_dim()` supplies the matching 3D space, which is what keeps the result
aligned with its source. The reduction itself is a lag-1 autocorrelation — one
number per voxel — and any vector-to-scalar function works in its place.

`vectors()` is the iterator form of the same idea, yielding one voxel time
course at a time:

```{r vectors}
length(vectors(bold))
```

## Putting objects together

`concat()` stacks along time. Volumes become a series:

```{r concat-vols}
dim(concat(anat, anat, anat))
```

and series extend each other, which is how runs get joined:

```{r concat-vecs}
run1 <- sub_vector(bold, 1:5)
run2 <- sub_vector(bold, 6:12)

dim(concat(run1, run2))
```

`concat()` takes the **first** argument's `NeuroSpace` for the result and does
not check that the others agree — a mismatched run is silently absorbed rather
than rejected. Check before you concatenate:

```{r concat-check}
identical(space(run1), space(run2))
```

`split_blocks()` is the inverse, cutting a concatenated series back into runs
given a block label per timepoint:

```{r split-blocks}
joined <- concat(run1, run2)
blocks <- split_blocks(joined, rep(1:2, c(5, 7)))

length(blocks)
vapply(blocks, function(b) dim(b)[4], integer(1))
```

`series()` pulls one voxel's time course directly, and `series_roi()` a whole
region's; regions, parcels and searchlights are the subject of
`vignette("regions-and-searchlights")`.

## Dense and sparse

When a mask defines which voxels count, a sparse representation stores only
those:

```{r sparse}
sparse <- as.sparse(bold, as.mask(mask))

class(sparse)[1]
dim(sparse)

c(dense_MB = as.numeric(object.size(bold)) / 1e6,
  sparse_MB = as.numeric(object.size(sparse)) / 1e6)
```

Dimensions are unchanged — sparsity is about storage, not shape — and
`as.dense()` reverses it. `read_vec(file, mask = ...)` reads straight into the
sparse form without materialising the dense one first.

```{r dense-again}
class(as.dense(sparse))[1]
```

When that trade is worth making, and the other backends available when data
outgrows memory, is the subject of `vignette("large-data")`.

## Where to go next

- `vignette("reading-and-writing")` — getting these objects to and from disk
- `vignette("regions-and-searchlights")` — ROIs, parcels, searchlights
- `vignette("large-data")` — sparse, mapped and file-backed storage

Reference: `?NeuroVol`, `?NeuroVec`, `?concat`, `?sub_vector`, `?as.mask`.
